---
title: "Differential Expression Analysis Visualization and Interpretation - Top Hits"
editor: source
author: "Joe Boktor"
date: '2026-01-01'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

# Goal: Examine the top differentially expressed genes regardless of cell type or tissue and identify any patterns or trends.

## Analysis

Loading environment and libraries

```{r}
library(Seurat)
library(Matrix)
library(irlba)
library(scico)
library(tidyverse)
library(patchwork)
library(glue)
library(presto)
library(hdf5r)
library(tidyseurat)
library(Augur)
library(batchtools)
library(doParallel)
library(foreach)


library(tictoc)
library(future)
library(strex)
library(cowplot)
library(aplot)
library(SingleCellExperiment)
library(BiocParallel)
library(ggside)
library(rstatix)
library(ggrepel)
library(ggpackets)

theme_set(theme_bw())

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
rnk_dir <- "/resnick/groups/mthomson/jboktor"
split_pipe_dir <- glue("{rnk_dir}/split-pipe_workflow")
wkdir <- glue("{home_dir}/WILDRxSPF_brains")
anndata_path <- glue("{wkdir}/data/input/parse_bio_anndata")
data_path <- glue("{wkdir}/data/interim")
fig_path <- glue("{wkdir}/figures")
abca_color_pal <- readRDS(glue("{wkdir}/data/interim/abca_color_pal.rds"))

source(glue("{wkdir}/notebooks/R_scripts/_misc_functions.R"))

# Set memory limit for future
options(future.globals.maxSize = 200000 * 1024^2)  # Set to ~200GB
# Use more memory-efficient serialization
options(future.serialization.progressr = FALSE)
options(future.serialization.progressr.enable = FALSE)

# future::plan(multisession, workers = 3)
# future::plan(sequential)


```


Loading Seurat object

```{r}
seurat_obj <- readRDS( 
    glue("{data_path}/seurat/snRNASeq/",
    "seurat_obj_onlyparsefilters_celltyped_2025-12-27.rds")
    )
seurat_obj
# Subset the seurat object to only keep cells that pass the QC filters
seurat_obj <- seurat_obj %>% subset(subset = keep_cell)
seurat_obj@meta.data %>% glimpse()
seurat_obj@meta.data$supertype_name %>% unique()

allen_annot_df <- seurat_obj@meta.data %>% 
  select(class_name, class_label, subclass_name, 
    subclass_label, supertype_name, supertype_label) %>% 
  distinct() %>% 
  glimpse()

allen_annot_long <- allen_annot_df %>% 
  pivot_longer(
    cols = everything(),
    names_to = c("cell_type_level", ".value"),
    names_pattern = "(class|subclass|supertype)_(name|label)"
  ) %>% 
  dplyr::rename(cell_type_name = name, cell_type_label = label) %>% 
  distinct() %>% 
  glimpse()

```


Loading and formatting MAST analysis results

```{r}
# clean  up format of MAST results
format_mastdf <- function(mast_df, name_col, level_col) {
  mast_df %>% 
    dplyr::rename(cell_type_name = {{name_col}}) %>% 
    janitor::clean_names() %>% 
    mutate(cell_type_level = {{level_col}}) %>% 
    glimpse()
}

# MAST Results (signficant hits only)
class_mast_df <- readRDS(glue("{data_path}/mast/snRNASeq/class_mast_signif_hits_2026-01-04.rds")) %>% 
    format_mastdf(name_col = "class_name", level_col = "class")
subclass_mast_df <- readRDS(
  glue("{data_path}/mast/snRNASeq/subclass_mast_signif_hits_2026-01-04.rds")) %>% 
    format_mastdf(name_col = "subclass_name", level_col = "subclass")
supertype_mast_df <- readRDS(
  glue("{data_path}/mast/snRNASeq/supertype_mast_signif_hits_2026-01-04.rds")) %>% 
    format_mastdf(name_col = "supertype_name", level_col = "supertype")
mast_df <- bind_rows(class_mast_df, subclass_mast_df, supertype_mast_df) %>% 
  filter(p_adjust < 0.05) %>% 
  mutate(gene = gsub("-GRCm39", "", primerid)) %>%
  mutate(direction = if_else(coef > 0, "WildR-UP", "SPF-UP")) %>%
  glimpse()

```


## Critical Questions to Answer

1. Which cell types have the most differentially expressed genes and are the most altered?

2. Which genes are altered most robustly? Are these highly significant genes altered in cell types that have the most alterations or from cell types that have just a few alterations?

3. Are there any tissue-specific signals?

4. Given the cell types with the largest alterations and the genes altered in these cell types, what pathways or circuits are most likely to be affected?

```{r}
celltype_droplist <- readRDS(
  glue("{data_path}/seurat/snRNASeq/celltype_droplist_2025-12-30.rds")
  )

metadata_df <- seurat_obj@meta.data %>% glimpse()
class_batchdata <- 
    dplyr::select(metadata_df, celltype = class_label, tissue) %>% 
    distinct() %>%
    mutate(cell_type_column = "class_label")
subclass_batchdata <- 
    dplyr::select(metadata_df, celltype = subclass_label, tissue) %>% 
    distinct() %>%
    mutate(cell_type_column = "subclass_label")
supertype_batchdata <- 
    dplyr::select(metadata_df, celltype = supertype_label, tissue) %>% 
    distinct() %>%
    mutate(cell_type_column = "supertype_label")

batchparams_df <- 
    bind_rows(class_batchdata, subclass_batchdata, supertype_batchdata) %>%
    left_join(celltype_droplist) %>%
    filter(is.na(dropme)) %>% 
    dplyr::select(cell_type_column, celltype, tissue) %>% 
    glimpse()

# number of supertypes tested in AMY and HYP
n_tested_AMY_supt <- batchparams_df %>% 
  filter(tissue == "AMY") %>% 
  filter(cell_type_column == "supertype_label") %>% 
  n_distinct()
n_tested_HYP_supt <- batchparams_df %>% 
  filter(tissue == "HYP") %>% 
  filter(cell_type_column == "supertype_label") %>% 
  n_distinct()

# number of supertypes with at least 1 DEG in AMY and HYP
mast_df %>% glimpse()
mast_df %>%
  filter(p_adjust < 0.05, abs(coef) > 0.25) %>% 
  filter(cell_type_level == "supertype") %>% 
  group_by(tissue, cell_type_name) %>% 
  tally() %>% 
  filter(n > 0) %>% 
  group_by(tissue) %>% 
  tally()

mast_df %>%
  filter(p_adjust < 0.05, abs(coef) > 0.25) %>% 
  filter(cell_type_level == "supertype") %>% 
  group_by(tissue, cell_type_name) %>% 
  tally() %>% 
  filter(n > 9) %>% 
  group_by(tissue) %>% 
  tally()

```


164 tested supertypes in AMY, 225 tested supertypes in HYP.

- Cell types with 1 or more DEGs in **AMY**: 51 (51/164 = 31%)

- Cell types with 1 or more DEGs in **HYP**: 124 (124/225 = 55%)

- Cell types with 10 or more DEGs in **AMY**: 26 (26/164 = 16%)

- Cell types with 10 or more DEGs in **HYP**: 9 (9/225 = 4%)



Do the most significant DEGs appear in the most altered cell types?

```{r}
most_significant_degs <- mast_df %>% 
  filter(p_adjust < 0.05) %>% 
  filter(abs(coef) > 0.25) %>% 
  arrange(p_adjust) %>% 
  pull(gene) %>% 
  unique() %>% 
  head(75)

most_sig_supt <- mast_df %>%
  filter(p_adjust < 0.05, abs(coef) > 0.25) %>% 
  filter(cell_type_level == "supertype") %>% 
  group_by(celltype_tissue) %>% 
  tally() %>% 
  filter(n > 9) %>% 
  pull(celltype_tissue)




```



## Conclusions / Key Findings

1. Which cell types have the most differentially expressed genes and are the most altered?

- Out of 225 supertypes tested in the HYP, 124 or (55%) have at least 1 DEG, and 9 or (4%) have at least 10 DEGs.

- Out of 164 supertypes tested in the AMY, 51 or (31%) have at least 1 DEG, and 26 or (16%) have at least 10 DEGs.

- A greater fraction of tested cells have at least 1 DEG in the HYP tissue than the AMY tissue; however, the fraction of cells with 10 or more DEGs is lower in the HYP tissue (4% vs 16%).



2. Which genes are altered most robustly? Are these highly significant genes altered in cell types that have the most alterations or from cell types that have just a few alterations?

- When we look at the top 75 DEGs, they are within X% of the most altered cell types. .. Here is important, are the highest DEGs also outside of these critical cell types?


3. Are there any tissue-specific signals?

- Yes, there are tissue-specific signals.

- X tissue has more cell types with robust alterations than Y tissue, 

- Some DEGs that sweep across multiple cell types in a tissue


4. Given the cell types with the largest alterations and the genes altered in these cell types, what pathways or circuits are most likely to be affected?






Visualizing a Datatable of top gene hits

```{r}
mast_df %>% 
  filter(cell_type_level == "supertype") %>% 
  arrange(p_adjust) %>% 
  dplyr::relocate(cell_type_name, 1) %>% 
  dplyr::relocate(gene, 1) %>% 
  select(-c(primerid, pr_chisq, ci_hi, ci_lo)) %>% 
  head(200) %>% 
  DT::datatable()

# visualizing the most significant p-value for each gene across all supertypes to identify top hits regardless of cell type or tissue
mast_df %>% 
  filter(cell_type_level == "supertype") %>%
  group_by(gene, direction, tissue) %>%
  summarize(min_p = min(p_adjust), n_signif = n()) %>% 
  arrange(min_p) %>% 
  mutate(min_p = formatC(min_p, format = "e", digits = 2)) %>% 
  head(200) %>%
  DT::datatable()

```


```{r}
# Top 100 supertype hits: prepare for dot-plot heatmap
top_mast_df <- mast_df %>% 
  filter(cell_type_level == "supertype") %>% 
  arrange(p_adjust) %>% 
  head(200) %>% 
  dplyr::mutate(
    gene = gsub("-GRCm39", "", primerid),
    neglog10_p = -log10(p_adjust)
  )

n_genes <- n_distinct(top_mast_df$gene)
n_supertypes <- n_distinct(top_mast_df$cell_type_name)
message(glue("Top 200 rows: {n_genes} unique genes across {n_supertypes} supertypes"))

# selecting the top 75 unique most significant genes for visualization
top_75_genes <- mast_df %>% 
  filter(cell_type_level == "supertype") %>% 
  arrange(p_adjust) %>% 
  pull(primerid) %>% 
  gsub("-GRCm39", "", .) %>% 
  unique() %>% 
  head(75)

# identifying cell types that have significance for any of these top 75 genes
significant_celltypes <- mast_df %>% 
  filter(cell_type_level == "supertype") %>% 
  filter(gsub("-GRCm39", "", primerid) %in% top_75_genes) %>% 
  group_by(cell_type_name) %>%
  tally() %>% 
  filter(n > 1) %>%
  pull(cell_type_name) %>% 
  unique()

message(glue("Top 75 unique genes are significant in {length(significant_celltypes)} supertypes"))

```

```{r}
# Dot-plot heatmap: genes vs supertypes
# Dots colored by coefficient (direction), sized by significance
p_dotplot_tophits <- mast_df %>% 
  filter(primerid %in% paste0(top_75_genes, "-GRCm39")) %>%
  filter(cell_type_name %in% significant_celltypes) %>% 
  dplyr::mutate(
    gene = gsub("-GRCm39", "", primerid),
    neglog10_p = -log10(p_adjust)
  ) %>% 
  ggplot(aes(
    x = fct_reorder(cell_type_name, cell_type_name, .fun = length),
    y = fct_reorder(gene, -neglog10_p, .fun = max),
    size = neglog10_p
  )) +
  geom_point(alpha = 0.9, shape = 21, aes(fill = coef)) +
  scale_fill_scico(
    palette = "cork",
    midpoint = 0,
    limits = c(-max(abs(top_mast_df$coef)), max(abs(top_mast_df$coef)))
  ) +
  scale_size_continuous(
    range = c(2, 8),
    name = expression(-log[10](p[adj]))
  ) +
  facet_grid(~ tissue, scales = "free", space = "free") +
  labs(
    x = "Supertype",
    y = "Gene",
    fill = "log2 FC",
    title = glue("Top 75 DEG hits: {length(top_75_genes)} genes × {length(significant_celltypes)} supertypes")
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 7, face = "italic"),
    panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 9, color = "grey30")
  )

ggsave(
    glue("{fig_path}/DEG/snRNASeq/top_hits/dotplot_top200_hits_{Sys.Date()}.png"),
    p_dotplot_tophits,
    width = 20,
    height = 10
)

```


```{r}
library(seriation)  # optional but has better dist handling for sparse matrices

# --- 1. Build the plotting dataframe first ---
plot_df <- mast_df %>%
  filter(primerid %in% paste0(top_75_genes, "-GRCm39")) %>%
  filter(cell_type_name %in% significant_celltypes) %>%
  dplyr::mutate(
    gene        = gsub("-GRCm39", "", primerid),
    neglog10_p  = -log10(p_adjust)
  )

# --- 2. Pivot to matrix for clustering (genes x celltypes) ---
# Use coef as the value; fill missing combos with 0
mast_df %>% glimpse()
clust_mat <- plot_df %>%
  dplyr::select(gene, celltype_tissue, coef) %>%
  tidyr::pivot_wider(
    names_from  = celltype_tissue,
    values_from = coef,
    values_fill = 0
  ) %>%
  tibble::column_to_rownames("gene") %>%
  as.matrix()

# --- 3. Cluster rows (genes) and columns (cell types) ---
row_order <- hclust(dist(clust_mat),            method = "ward.D2")$order
col_order <- hclust(dist(t(clust_mat)),         method = "ward.D2")$order

gene_levels     <- rownames(clust_mat)[row_order]
celltype_levels <- colnames(clust_mat)[col_order]
# --- 4. Apply clustered factor levels ---
plot_df <- plot_df %>%
  mutate(
    gene            = factor(gene,            levels = gene_levels),
    celltype_tissue = factor(celltype_tissue, levels = celltype_levels)
  )

# --- 4b. Build the label lookup (celltype_tissue -> cell_type_name) ---
ct_labels <- plot_df %>%
  distinct(celltype_tissue, cell_type_name) %>%
  deframe()  # named vector: names = celltype_tissue, values = cell_type_name

# --- 5. Plot ---
p_dotplot_tophits <- plot_df %>%
  ggplot(aes(
    x    = celltype_tissue,
    y    = gene,
    size = neglog10_p
  )) +
  geom_point(alpha = 0.9, shape = 21, aes(fill = coef)) +
  scale_x_discrete(labels = ct_labels) +          # <-- swap display labels here
  scale_fill_scico(
    palette  = "cork",
    midpoint = 0,
    limits   = c(-max(abs(top_mast_df$coef)), max(abs(top_mast_df$coef)))
  ) +
  scale_size_continuous(
    range = c(2, 8),
    name  = expression(-log[10](p[adj]))
  ) +
  facet_grid(~ tissue, scales = "free", space = "free") +
  labs(
    x     = "Supertype",
    y     = "Gene",
    fill  = "log2 FC",
    title = glue("Top 75 DEG hits: {length(top_75_genes)} genes × {length(significant_celltypes)} supertypes")
  ) +
  theme_bw() +
  theme(
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y   = element_text(size = 7, face = "italic"),
    panel.grid.major = element_line(linewidth = 0.2, color = "grey90"),
    legend.position  = "right",
    plot.title    = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 9, color = "grey30")
  )

ggsave(
    glue("{fig_path}/DEG/snRNASeq/top_hits/dotplot_top200_hits_{Sys.Date()}.png"),
    p_dotplot_tophits,
    width = 18,
    height = 10
)

```


Looking at P-value distribution of top hits

```{r}

mast_df %>% 
  filter(primerid %in% paste0(top_75_genes, "-GRCm39")) %>%
  filter(cell_type_name %in% significant_celltypes) %>% 
  dplyr::mutate(
    gene = gsub("-GRCm39", "", primerid),
    neglog10_p = -log10(p_adjust)
  ) %>% 
  filter(gene %in% c("Cmss1", "Cntn6") ) %>% 
  ggplot(aes(x=neglog10_p)) +
  # geom_density(alpha = 0.5, aes(fill = gene))
  geom_histogram(position = "identity", alpha = 0.6, bins = 30, aes(fill = gene))

```



Plot violin plots for top hits

```{r}
# Hits to plot: top 300 most significant OR genes in cell-type+tissue combos with >= 5 hits
# Output: {cell_type_level}/{tissue}/{WildR_UP|SPF_UP}/{gene}_{tissue}_{celltype}_{date}.png
cell_type_level <- "supertype"
celltype_column <- "supertype_name"
top_n <- 300
min_hits_per_celltype_tissue <- 5

supertype_mast <- mast_df %>%
  filter(cell_type_level == !!cell_type_level) %>%
  filter(p_adjust <= 0.05) %>%
  filter(abs(coef) > 0.25)

# Cell-type+tissue combos with >= min_hits_per_celltype_tissue hits
high_hit_combos <- supertype_mast %>%
  group_by(cell_type_name, tissue) %>%
  tally() %>%
  filter(n >= min_hits_per_celltype_tissue) %>%
  select(cell_type_name, tissue)

# Top top_n by p_adjust
top_by_p <- supertype_mast %>%
  arrange(p_adjust) %>%
  head(top_n) %>%
  select(gene, cell_type_name, tissue, direction)

# Genes in high-hit cell types (all their hits, not just top 300)
high_hit_genes <- supertype_mast %>%
  semi_join(high_hit_combos, by = c("cell_type_name", "tissue")) %>%
  select(gene, cell_type_name, tissue, direction)

# Union: plot if in top 300 OR in a high-hit cell type
violin_hits <- bind_rows(top_by_p, high_hit_genes) %>%
  distinct(gene, cell_type_name, tissue, direction) %>% 
  mutate(cell_type_name_clean = gsub("/", "-", cell_type_name) %>% gsub(" ", "_", .)) %>% 
  glimpse()

message(glue("Plotting {nrow(violin_hits)} violin plots"))

```


```{r}
# Fetch expression once per (celltype, tissue), then loop through plots
violin_out_base <- glue("{fig_path}/DEG/snRNASeq/top_hits/{cell_type_level}")

violin_hits_grouped <- violin_hits %>%
    # temporarily fix for Pakap-2 issue
    filter(gene %nin% c("Pakap-2")) %>% 
    group_by(cell_type_name, cell_type_name_clean, tissue) %>%
    summarise(genes = list(tibble(gene, direction)), .groups = "drop")

# cl <- makeCluster(parallel::detectCores() - 1)
registerDoParallel(7)

for (g in seq_len(nrow(violin_hits_grouped))) {
  celltype_index <- violin_hits_grouped$cell_type_name[g]
  celltype        <- violin_hits_grouped$cell_type_name_clean[g]
  tissue          <- violin_hits_grouped$tissue[g]
  gene_dir        <- violin_hits_grouped$genes[[g]]
  genes           <- gene_dir$gene

  multi_df <- get_gene_expression_df_multi(
    seurat_obj,
    celltype_column = celltype_column,
    celltype_value  = celltype_index,
    genes           = genes,
    tissue          = tissue,
    return_format   = "log2cpm"
  ) %>% glimpse()

  if (nrow(multi_df) == 0) {
    warning(glue("No cells for {celltype} / {tissue}; skipping {length(genes)} genes"))
    next
  }

  multi_df <- multi_df %>%
    mutate(sample = factor(sample, levels = unique(sample[order(microbiome, sample)])))

  # Capture loop-level variables for export into workers
  .multi_df  <- multi_df
  .gene_dir  <- gene_dir
  .tissue    <- tissue
  .celltype  <- celltype
  .out_base  <- violin_out_base

  foreach(
    j = seq_len(nrow(.gene_dir)),
    .packages  = c("ggplot2", "glue", "dplyr"),   # add any others your fns need
    .export    = c(".multi_df", ".gene_dir", ".tissue", ".celltype", ".out_base",
                   "plot_gene_expression")
  ) %dopar% {
    genename  <- .gene_dir$gene[j]
    direction <- .gene_dir$direction[j]
    out_dir   <- glue("{.out_base}/{.tissue}/{.celltype}")
    fname     <- glue("{genename}_{direction}_{.tissue}_{.celltype}.png")
    full_fig_path <- file.path(out_dir, fname)

    if (file.exists(full_fig_path)) return(NULL)   # skip instead of next
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

    p_combined <- plot_gene_expression(
      .multi_df,
      gene           = genename,
      tissue         = .tissue,
      celltype_value = .celltype
    )
    ggsave(full_fig_path, p_combined, width = 8, height = 5)
    NULL
  }
}

stopCluster(cl)
```
