---
title: "Differential Expression Analysis Visualization and Interpretation (Cmss1-Lars)"
editor: source
author: "Joe Boktor"
date: '2026-08-09'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

```{r}
library(Seurat)
library(Matrix)
library(irlba)
library(scico)
library(tidyverse)
library(patchwork)
library(glue)
library(presto)
library(hdf5r)
library(tidyseurat)
library(Augur)
library(batchtools)
library(ggpubr)

library(tictoc)
library(future)
library(strex)
library(cowplot)
library(aplot)
library(SingleCellExperiment)
library(BiocParallel)
library(ggside)
library(rstatix)
library(ggrepel)
library(ggpackets)

theme_set(theme_bw())

home_dir <- "/resnick/groups/mthomson/jboktor"
rnk_dir <- "/resnick/groups/mthomson/jboktor"
split_pipe_dir <- glue("{rnk_dir}/split-pipe_workflow")
wkdir <- glue("{home_dir}/WILDRxSPF_brains")
anndata_path <- glue("{wkdir}/data/input/parse_bio_anndata")
data_path <- glue("{wkdir}/data/interim")
fig_path <- glue("{wkdir}/figures")
abca_color_pal <- readRDS(glue("{wkdir}/data/interim/abca_color_pal.rds"))

source(glue("{wkdir}/notebooks/R_scripts/_misc_functions.R"))

# Set memory limit for future
options(future.globals.maxSize = 200000 * 1024^2)  # Set to ~200GB
# Use more memory-efficient serialization
options(future.serialization.progressr = FALSE)
options(future.serialization.progressr.enable = FALSE)

# future::plan(multisession, workers = 3)
# future::plan(sequential)

```


Loading Seurat object


```{r}
seurat_obj <- readRDS( 
    glue("{data_path}/seurat/snRNASeq/",
    "seurat_obj_onlyparsefilters_celltyped_2025-12-27.rds")
    )
seurat_obj
# Subset the seurat object to only keep cells that pass the QC filters
seurat_obj <- seurat_obj %>% subset(subset = keep_cell)
seurat_obj@meta.data %>% glimpse()
seurat_obj@meta.data$supertype_name %>% unique()

allen_annot_df <- seurat_obj@meta.data %>% 
  select(class_name, class_label, subclass_name, 
    subclass_label, supertype_name, supertype_label) %>% 
  distinct() %>% 
  glimpse()

allen_annot_long <- allen_annot_df %>% 
  pivot_longer(
    cols = everything(),
    names_to = c("cell_type_level", ".value"),
    names_pattern = "(class|subclass|supertype)_(name|label)"
  ) %>% 
  dplyr::rename(cell_type_name = name, cell_type_label = label) %>% 
  distinct() %>%
  glimpse()

```


## Does Cmss1/Lars2 expression correlate with unique UMIs per cell?

Look at a handful of cell types in amygdala (AMY) and hypothalamus (HYP). One plot per
supertype; AMY and HYP share the same plot as tissue facets, with Pearson correlation
computed within each tissue split.


```{r}
#| label: cmss1-lars2-umi-cor-by-supertype

supertypes_of_interest <- c(
  "1160 Astro-NT NN_2",
  "1179 OPC NN_1",
  "0631 MM Foxb1 Glut_2"
)
genes_of_interest <- c("Cmss1", "Lars2")
tissues_of_interest <- c("AMY", "HYP")

# Confirm exact supertype strings in the object
supt <- unique(seurat_obj@meta.data$supertype_name)
walk(c("1160", "1179", "0631"), \(id) print(supt[grep(id, supt)]))

cmss1_lars_umi_df <- map_dfr(
  supertypes_of_interest,
  \(stype) {
    get_gene_expression_df_multi(
      seurat_obj,
      celltype_column = "supertype_name",
      celltype_value = stype,
      genes = genes_of_interest,
      tissue = NULL,
      return_format = "log2cpm"
    ) %>%
      as_tibble(rownames = "cell_id") %>%
      filter(tissue %in% tissues_of_interest) %>%
      mutate(
        tissue = factor(tissue, levels = tissues_of_interest),
        supertype_name = stype
      )
  }
) %>% 
  glimpse()

cmss1_lars_umi_long <- cmss1_lars_umi_df %>%
  pivot_longer(
    cols = all_of(paste0(genes_of_interest, "_log2cpm")),
    names_to = "gene",
    values_to = "gene_log2cpm",
    names_transform = list(gene = \(x) str_remove(x, "_log2cpm$"))
  ) %>%
  mutate(gene = factor(gene, levels = genes_of_interest)) %>% 
  glimpse()

# Stratified Pearson correlations: gene × tissue × supertype
cmss1_lars_umi_cors <- cmss1_lars_umi_long %>%
  group_by(supertype_name, tissue, gene) %>%
  summarise(
    n_cells = n(),
    pearson_r = cor(nCount_RNA, gene_log2cpm, use = "complete.obs", method = "pearson"),
    spearman_rho = cor(nCount_RNA, gene_log2cpm, use = "complete.obs", method = "spearman"),
    p_pearson = tryCatch(
      cor.test(nCount_RNA, gene_log2cpm, method = "pearson")$p.value,
      error = \(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  arrange(supertype_name, gene, tissue) %>% 
  glimpse()

readr::write_csv(cmss1_lars_umi_long,
  glue("{data_path}/misc/cmss1_lars2_umi_long_{Sys.Date()}.csv")
  )

print(cmss1_lars_umi_cors)

# One plot per supertype; AMY / HYP as columns, Cmss1 / Lars2 as rows
p_cmss1_lars_umi_list <- map(
  supertypes_of_interest,
  \(stype) {
    df_i <- filter(cmss1_lars_umi_long, supertype_name == stype)
    ggplot(df_i, aes(x = nCount_RNA, y = gene_log2cpm)) +
      geom_point(aes(color = tissue), alpha = 0.25, size = 0.6) +
      geom_smooth(method = "lm", se = TRUE, linewidth = 0.7, color = "grey20") +
      ggpubr::stat_cor(
        method = "pearson",
        label.x.npc = 0.05,
        label.y.npc = 0.97,
        size = 3
      ) +
      facet_grid(gene ~ tissue, scales = "free") +
      scale_x_log10() +
      scale_color_manual(values = c(AMY = "#54A24B", HYP = "#B279A2")) +
      theme_bw() +
      theme(legend.position = "none") +
      labs(
        title = stype,
        subtitle = "Cmss1 / Lars2 log2CPM vs unique UMIs (nCount_RNA), stratified by tissue",
        x = "Unique UMIs per cell (nCount_RNA)",
        y = "Gene expression (log2CPM)"
      )
  }
)
names(p_cmss1_lars_umi_list) <- supertypes_of_interest

walk(p_cmss1_lars_umi_list, print)

p_cmss1_lars_umi_all <- wrap_plots(p_cmss1_lars_umi_list, ncol = 1)
p_cmss1_lars_umi_all

dir.create(glue("{fig_path}/DEG/snRNASeq/misc"), recursive = TRUE, showWarnings = FALSE)
ggsave(
  glue("{fig_path}/DEG/snRNASeq/misc/cmss1_lars2_umi_cor_by_supertype_{Sys.Date()}.png"),
  p_cmss1_lars_umi_all,
  width = 10,
  height = 14,
  dpi = 300
)

```


## Ambient / contamination gene raw counts

Export raw UMI counts (not log2CPM) for classic ambient genes in the same
supertypes × AMY/HYP cells as above. Gene inventory:

- Present: `Rn18s-rs5`, `Malat1`, `Erh`, all `mt-*`, `Rpl*` / `Rps*` (excl. `Rps6k*` kinases)
- Absent from this GRCm39 object (and source `wildrxspf_data.h5ad`): `Gm42418`, `AY036118`
  (`Gm42418` is the mm10 alias of `Rn18s-rs5`)


```{r}
#| label: ambient-contam-genes-raw-counts-export

feats <- rownames(seurat_obj)
feats_clean <- strex::str_before_first(feats, "-GRCm39")
feats_clean[is.na(feats_clean)] <- feats[is.na(feats_clean)]

exact_genes <- c("Rn18s-rs5", "Gm42418", "AY036118", "Malat1", "Erh")
exact_present <- exact_genes[exact_genes %in% feats_clean]
exact_missing <- setdiff(exact_genes, exact_present)
mt_genes <- sort(unique(feats_clean[startsWith(feats_clean, "mt-")]))
rpl_rps_genes <- sort(unique(feats_clean[
  (startsWith(feats_clean, "Rpl") | startsWith(feats_clean, "Rps")) &
    !startsWith(feats_clean, "Rps6k")
]))
ambient_genes <- unique(c(exact_present, mt_genes, rpl_rps_genes))

message("Ambient genes present: ", length(ambient_genes))
message("Exact missing: ", paste(exact_missing, collapse = ", "))

ambient_raw_df <- map_dfr(
  supertypes_of_interest,
  \(stype) {
    get_gene_expression_df_multi(
      seurat_obj,
      celltype_column = "supertype_name",
      celltype_value = stype,
      genes = ambient_genes,
      tissue = NULL,
      return_format = "raw"
    ) %>%
      as_tibble(rownames = "cell_id") %>%
      filter(tissue %in% tissues_of_interest) %>%
      mutate(
        tissue = factor(tissue, levels = tissues_of_interest),
        supertype_name = stype
      )
  }
)

ambient_raw_long <- ambient_raw_df %>%
  pivot_longer(
    cols = all_of(paste0(ambient_genes, "_raw")),
    names_to = "gene",
    values_to = "gene_raw",
    names_transform = list(gene = \(x) str_remove(x, "_raw$"))
  ) %>%
  mutate(gene = factor(gene, levels = ambient_genes)) %>%
  glimpse()

dir.create(glue("{data_path}/misc"), recursive = TRUE, showWarnings = FALSE)
readr::write_csv(
  ambient_raw_long,
  glue("{data_path}/misc/ambient_contam_genes_raw_counts_long_{Sys.Date()}.csv")
)
message(
  "Wrote ",
  glue("{data_path}/misc/ambient_contam_genes_raw_counts_long_{Sys.Date()}.csv")
)

```


Running scCDC contamination detection on HYP cells

```{r, eval = FALSE}
library(scCDC)
seur_hyp <- subset(seurat_obj, subset = tissue == "HYP")
gcgs <- ContaminationDetection(seur_hyp)
# > gcgs <- ContaminationDetection(seur_hyp)
# Caculating entropy...
#   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=12m 54s
# Caculating expression level...
#   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=09m 08s
# Calculating entropy-expression relation...
#   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=35s  
# Extracting contamination degree...
#   |++++++++++++++++++++++++++++++++++++++++++++++++++| 100% elapsed=00s  
# Complete detection. 340 contaminated genes found
# There were 50 or more warnings (use warnings() to see the first 50)
# > 
saveRDS(gcgs, glue("{data_path}/misc/hyp_gcgs_{Sys.Date()}.rds"))

seurat_obj@meta.data %>% glimpse()

```


Searching for Cmss1 / Lars2 in the scCDC contamination detection results. 

```{r}
#| label: searching-for-cmss1-lars2-in-gcgs

gcgs <- readRDS(glue("{data_path}/misc/hyp_gcgs_2026-08-09.rds"))
gcgs %>% glimpse()
gcgs %>% str()
gcgs %>% View()

gcgs[grepl("Cmss1", rownames(gcgs))]
gcgs[grepl("Lars2", rownames(gcgs))]
gcgs[grepl("95041", rownames(gcgs))]

```


Cmss1 and Lars2 are not present in the scCDC contamination detection results. 


** Data to be convinced this is not a contaminant/artifact ** 

- Does this finding hold with sample labels shuffled MannWhitney U tests? (permutation test) - Nominally
- Is there a correlation with known ambient / contamination genes? - None
- Look for detection of Cmss1 or Lars2 in scCDC contamination detection results. - Not present
- Correlation between UMI counts and gene expression - None found


Okay moving on, assuming this is a soft but possibly interesting signal - what could be driving this signature and what additional test would enlighten its role?


Looking at correlations of Cmss1 / Lars2 expression with scFEA single cell flux modules for 

```{r}

```

