---
title: "Differential Expression Analysis Visualization and Interpretation"
editor: source
author: "Joe Boktor"
date: '2026-01-01'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

```{r}
library(Seurat)
library(Matrix)
library(irlba)
library(scico)
library(tidyverse)
library(patchwork)
library(glue)
library(presto)
library(hdf5r)
library(tidyseurat)
library(Augur)
library(batchtools)

library(tictoc)
library(future)
library(strex)
library(cowplot)
library(aplot)
library(SingleCellExperiment)
library(BiocParallel)
library(ggside)
library(rstatix)
library(ggrepel)
library(ggpackets)

theme_set(theme_bw())

home_dir <- "/resnick/groups/mthomson/jboktor"
rnk_dir <- "/resnick/groups/mthomson/jboktor"
split_pipe_dir <- glue("{rnk_dir}/split-pipe_workflow")
wkdir <- glue("{home_dir}/WILDRxSPF_brains")
anndata_path <- glue("{wkdir}/data/input/parse_bio_anndata")
data_path <- glue("{wkdir}/data/interim")
fig_path <- glue("{wkdir}/figures")
abca_color_pal <- readRDS(glue("{wkdir}/data/interim/abca_color_pal.rds"))

source(glue("{wkdir}/notebooks/R_scripts/_misc_functions.R"))

# Set memory limit for future
options(future.globals.maxSize = 200000 * 1024^2)  # Set to ~200GB
# Use more memory-efficient serialization
options(future.serialization.progressr = FALSE)
options(future.serialization.progressr.enable = FALSE)

# future::plan(multisession, workers = 3)
# future::plan(sequential)

```


Loading Seurat object


```{r}
seurat_obj <- readRDS( 
    glue("{data_path}/seurat/snRNASeq/",
    "seurat_obj_onlyparsefilters_celltyped_2025-12-27.rds")
    )
seurat_obj
# Subset the seurat object to only keep cells that pass the QC filters
seurat_obj <- seurat_obj %>% subset(subset = keep_cell)
seurat_obj@meta.data %>% glimpse()
seurat_obj@meta.data$supertype_name %>% unique()

allen_annot_df <- seurat_obj@meta.data %>% 
  select(class_name, class_label, subclass_name, 
    subclass_label, supertype_name, supertype_label) %>% 
  distinct() %>% 
  glimpse()

allen_annot_long <- allen_annot_df %>% 
  pivot_longer(
    cols = everything(),
    names_to = c("cell_type_level", ".value"),
    names_pattern = "(class|subclass|supertype)_(name|label)"
  ) %>% 
  dplyr::rename(cell_type_name = name, cell_type_label = label) %>% 
  distinct() %>% 
  glimpse()

```


Loading Analysis Results

```{r}
# clean  up format of MAST results
format_mastdf <- function(mast_df, name_col, level_col) {
  mast_df %>% 
    dplyr::rename(cell_type_name = {{name_col}}) %>% 
    janitor::clean_names() %>% 
    mutate(cell_type_level = {{level_col}}) %>% 
    glimpse()
}

# MAST Results (signficant hits only)
class_mast_df <- readRDS(glue("{data_path}/mast/snRNASeq/class_mast_signif_hits_2026-01-04.rds")) %>% 
    format_mastdf(name_col = "class_name", level_col = "class")
subclass_mast_df <- readRDS(
  glue("{data_path}/mast/snRNASeq/subclass_mast_signif_hits_2026-01-04.rds")) %>% 
    format_mastdf(name_col = "subclass_name", level_col = "subclass")
supertype_mast_df <- readRDS(
  glue("{data_path}/mast/snRNASeq/supertype_mast_signif_hits_2026-01-04.rds")) %>% 
    format_mastdf(name_col = "supertype_name", level_col = "supertype")
mast_df <- bind_rows(class_mast_df, subclass_mast_df, supertype_mast_df) %>% 
  filter(p_adjust < 0.1) %>% 
    glimpse()

write_csv(mast_df, glue("{data_path}/mast/snRNASeq/mast_signif_hits_2026-08-05_p_adjust_0.1.csv"))

# Augur Results
format_augurdf <- function(augur_df, name_col, level_col) {
  augur_df %>% 
    dplyr::rename(cell_type_name = {{name_col}}) %>% 
    mutate(cell_type_level = {{level_col}}) %>% 
    glimpse()
}

augur_resdir <- glue("{data_path}/augur/snRNASeq/var_thresh_0.6")
class_aucs_df <- readRDS(
  glue("{augur_resdir}/class_aucs_df_2026-01-10.rds")) %>% 
  format_augurdf(name_col = "class_name", level_col = "class")
subclass_aucs_df <- readRDS(
  glue("{data_path}/augur/snRNASeq/var_thresh_0.6/subclass_aucs_df_2026-01-10.rds")) %>% 
  format_augurdf(name_col = "subclass_name", level_col = "subclass")
supertype_aucs_df <- readRDS(
  glue("{data_path}/augur/snRNASeq/var_thresh_0.6/supertype_aucs_df_2026-01-10.rds")) %>% 
  format_augurdf(name_col = "supertype_name", level_col = "supertype")
augur_df <- bind_rows(class_aucs_df, subclass_aucs_df, supertype_aucs_df) %>% 
    glimpse()

# PseudoBulk Results (EdgeR LRT)
edgeR_res <- readRDS(
  glue("{data_path}/edgeR/snRNASeq/edgeR_signif_hits_lrt_res_2026-01-03.rds")) %>% 
  mutate(
    tissue = str_after_last(celltype_tissue, "_"),
    cell_type = str_before_last(celltype_tissue, "_"), 
    cell_type_label = cell_type) %>% 
  filter(FDR < 0.1) %>% 
  # filter(PValue < 0.05) %>% 
  left_join(allen_annot_long) %>% 
  glimpse()

# Deseq2 Results
deseq2_res <- readRDS(
  glue("{data_path}/DESeq2/snRNASeq/deseq2_signif_hits_res_2026-01-03.rds")) %>% 
  mutate(
    tissue = str_after_last(celltype_tissue, "_"), 
    cell_type = str_before_last(celltype_tissue, "_"), 
    cell_type_label = cell_type) %>% 
  left_join(allen_annot_long) %>% 
  filter(padj < 0.1) %>% 
  left_join(allen_annot_long) %>% 
  glimpse()

```


```{r}
mast_df %>% glimpse()
mast_df %>% View()


```



Making volcano plots for each top hit supertype tissue combination
```{r}
# getting a list of supertypes that have a minimum number of DEGs to make volcano plots for
deseq2_suptype_tiss_list <- deseq2_res %>%
  filter(cell_type_level == "supertype") %>%
  group_by(celltype_tissue) %>% 
  tally() %>% 
  filter(n > 3) %>% 
  pull(celltype_tissue) %>% 
  unique()
edgeR_suptype_tiss_list <-edgeR_res %>%
  filter(cell_type_level == "supertype") %>%
  group_by(celltype_tissue) %>% 
  tally() %>% 
  filter(n > 3) %>% 
  pull(celltype_tissue) %>% 
  unique()
mast_suptype_tiss_list <-mast_df %>%
  filter(cell_type_level == "supertype") %>%
  group_by(celltype_tissue) %>% 
  tally() %>% 
  filter(n > 5) %>% 
  pull(celltype_tissue) %>% 
  unique()
suptype_tiss_list <- list(
  deseq2 = deseq2_suptype_tiss_list,
  edgeR = edgeR_suptype_tiss_list,
  mast = mast_suptype_tiss_list
)

supertype_tissue_to_plot <-unlist(suptype_tiss_list) %>% unname() %>% unique()
supertype_map <- seurat_obj@meta.data %>% 
  dplyr::select(supertype_name, supertype_label) %>% 
  distinct() %>% 
  glimpse()

# supertype_tissue <- "CS20230722_SUPT_0626_HYP"
for (supertype_tissue in supertype_tissue_to_plot) {
  tissue <- str_after_last(supertype_tissue, "_")
  supertype_label_i <- str_before_last(supertype_tissue, "_")
  supertype <- supertype_map %>% filter(supertype_label == supertype_label_i) %>% pull(supertype_name)
  supertype_clean <- gsub("/", "_", supertype)
  
  deseq2_degdf <- readRDS(
    glue("{data_path}/DESeq2/snRNASeq/supertype_label/{supertype_tissue}/res_lfcShrink_2025-12-30.rds"))
  edgeR_degdf <- readRDS(
    glue("{data_path}/edgeR/snRNASeq/supertype_label/{supertype_tissue}/res_lrt_2025-12-30.rds"))
  mast_degdf <- readRDS(
    glue("{data_path}/mast/snRNASeq/supertype_label/{supertype_tissue}_2026-01-04.rds"))

  top_genes_deseq2 <- get_top_deg_genes(deseq2_degdf, n = 15, p_threshold = 0.05)
  top_genes_edgeR <- get_top_deg_genes(edgeR_degdf, n = 15, p_threshold = 0.05)
  top_genes_mast <- get_top_deg_genes(mast_degdf, n = 15, p_threshold = 0.05)

  # Volcano plot template (ggpackets) - shared layers for all three DEG tools
  deseq2_plot_df <- deseq2_degdf %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    mutate(
      gene_label = gsub("-GRCm39$", "", gene),
      neglog10padj = -log10(padj + 1e-300),
      sig = padj < 0.05 & abs(log2FoldChange) > 0.25
    )
  deseq2_labels <- deseq2_plot_df %>%
    filter(gene_label %in% c(top_genes_deseq2$up, top_genes_deseq2$down))
  p_dseq_volc <- deseq2_plot_df %>%
    ggplot(aes(x = log2FoldChange, y = neglog10padj, color = sig)) +
    ggpk_volcano_deg(deseq2_labels, "log2FoldChange") +
    labs(title = "DESeq2 Volcano Plot")

  edgeR_plot_df <- edgeR_degdf %>%
    as.data.frame() %>%
    rownames_to_column("gene") %>%
    mutate(
      gene_label = gsub("-GRCm39$", "", gene),
      neglog10padj = -log10(FDR + 1e-300),
      sig = FDR < 0.05 & abs(logFC) > 0.25
    )
  edgeR_labels <- edgeR_plot_df %>%
    filter(gene_label %in% c(top_genes_edgeR$up, top_genes_edgeR$down))
  p_edgeR_volc <- edgeR_plot_df %>%
    ggplot(aes(x = logFC, y = neglog10padj, color = sig)) +
    ggpk_volcano_deg(edgeR_labels, "logFC") +
    labs(title = "EdgeR Volcano Plot")

  mast_plot_df <- mast_degdf %>%
    dplyr::rename(gene = primerid) %>%
    mutate(
      gene_label = gsub("-GRCm39$", "", gene),
      neglog10padj = -log10(p_adjust + 1e-300),
      sig = p_adjust < 0.05 & abs(coef) > 0.25
    )
  mast_labels <- mast_plot_df %>%
    filter(gene_label %in% c(top_genes_mast$up, top_genes_mast$down))
  p_mast_volc <- mast_plot_df %>%
    ggplot(aes(x = coef, y = neglog10padj, color = sig)) +
    ggpk_volcano_deg(mast_labels, "coef") +
    labs(title = "MAST Volcano Plot")

  p_combined_volc <- plot_grid(p_mast_volc, p_dseq_volc, p_edgeR_volc, ncol = 3, labels = "AUTO")
  ggsave(
    glue("{fig_path}/DEG/snRNASeq/volcano_plots/{supertype_tissue}_{supertype_clean}_combined_volcano_plot_{Sys.Date()}.png"),
    p_combined_volc,
    width = 12,
    height = 4
  )

}

#---------------------------------------------------------
# Manaual example
tissue <- "HYP"
deseq2_degdf <- readRDS(
  glue("{data_path}/DESeq2/snRNASeq/supertype_label/CS20230722_SUPT_0626_{tissue}/res_lfcShrink_2025-12-30.rds"))
edgeR_degdf <- readRDS(
  glue("{data_path}/edgeR/snRNASeq/supertype_label/CS20230722_SUPT_0626_{tissue}/res_lrt_2025-12-30.rds"))
mast_degdf <- readRDS(
  glue("{data_path}/mast/snRNASeq/supertype_label/CS20230722_SUPT_0626_{tissue}_2026-01-04.rds"))

top_genes_deseq2 <- get_top_deg_genes(deseq2_degdf, n = 10, p_threshold = 0.05)
top_genes_edgeR <- get_top_deg_genes(edgeR_degdf, n = 10, p_threshold = 0.05)
top_genes_mast <- get_top_deg_genes(mast_degdf, n = 10, p_threshold = 0.05)


# Volcano plot template (ggpackets) - shared layers for all three DEG tools
deseq2_plot_df <- deseq2_degdf %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  mutate(
    gene_label = gsub("-GRCm39$", "", gene),
    neglog10padj = -log10(padj + 1e-300),
    sig = padj < 0.05 & abs(log2FoldChange) > 0.25
  )
deseq2_labels <- deseq2_plot_df %>%
  filter(gene_label %in% c(top_genes_deseq2$up, top_genes_deseq2$down))
p_dseq_volc <- deseq2_plot_df %>%
  ggplot(aes(x = log2FoldChange, y = neglog10padj, color = sig)) +
  ggpk_volcano_deg(deseq2_labels, "log2FoldChange") +
  labs(title = "DESeq2 Volcano Plot")

edgeR_plot_df <- edgeR_degdf %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  mutate(
    gene_label = gsub("-GRCm39$", "", gene),
    neglog10padj = -log10(FDR + 1e-300),
    sig = FDR < 0.05 & abs(logFC) > 0.25
  )
edgeR_labels <- edgeR_plot_df %>%
  filter(gene_label %in% c(top_genes_edgeR$up, top_genes_edgeR$down))
p_edgeR_volc <- edgeR_plot_df %>%
  ggplot(aes(x = logFC, y = neglog10padj, color = sig)) +
  ggpk_volcano_deg(edgeR_labels, "logFC") +
  labs(title = "EdgeR Volcano Plot")

mast_plot_df <- mast_degdf %>%
  dplyr::rename(gene = primerid) %>%
  mutate(
    gene_label = gsub("-GRCm39$", "", gene),
    neglog10padj = -log10(p_adjust + 1e-300),
    sig = p_adjust < 0.05 & abs(coef) > 0.25
  )
mast_labels <- mast_plot_df %>%
  filter(gene_label %in% c(top_genes_mast$up, top_genes_mast$down))
p_mast_volc <- mast_plot_df %>%
  ggplot(aes(x = coef, y = neglog10padj, color = sig)) +
  ggpk_volcano_deg(mast_labels, "coef") +
  labs(title = "MAST Volcano Plot")

p_combined_volc <- plot_grid(p_mast_volc, p_dseq_volc, p_edgeR_volc, ncol = 3, labels = "AUTO")
# p_combined_volc
ggsave(
  glue("{fig_path}/DEG/snRNASeq/volcano_plots/combined_volcano_plot_{Sys.Date()}.png"),
  p_combined_volc,
  width = 12,
  height = 4
)

```

Searching for speciic genes of interest
```{r}
mast_df %>% glimpse()
names <- mast_df$cell_type_name %>% unique() 
names[which(grepl("1160", names))]

mast_df %>% 
  filter(cell_type_name == "0568 VMH Nr5a1 Glut_3") %>% 
  glimpse()

mast_wildrup <- mast_df %>% 
  filter(cell_type_name == "0568 VMH Nr5a1 Glut_3") %>%
  filter(coef > 0) %>% 
  filter(p_adjust < 0.01) %>%
  arrange(p_adjust) %>% 
  pull(primerid) %>% 
  gsub("-GRCm39", "", .)

mast_wildrdw <-mast_df %>% 
  filter(cell_type_name == "0568 VMH Nr5a1 Glut_3") %>%
  filter(coef < 0) %>% 
  filter(p_adjust < 0.01) %>% 
  arrange(p_adjust) %>% 
  pull(primerid) %>% 
  gsub("-GRCm39", "", .)

deseq2_res %>% 
  filter(cell_type_name == "0626 PH-SUM Foxa1 Glut_5") %>%
  filter(padj < 0.05) %>%
  arrange(padj) %>%
  pull(gene) %>%
  gsub("-GRCm39", "", .)

```


Ignoring how many DEGs a cell type has, what are the strongest hits?

```{r}
mast_df %>% filter(cell_type_level == "supertype") %>% 
  arrange(p_adjust) %>% 
  View
  group_by(celltype_tissue) %>% 
  tally() %>% 
  filter(n > 5) %>% 
  pull(celltype_tissue) %>% 
  unique()

```


Making gene violin plots for specific genes of interest from top cell hits

```{r}

# list of supertypes of interest
mast_df %>% glimpse()
mast_df %>%
  filter(cell_type_level == "supertype") %>% 
  group_by(celltype_tissue) %>% 
  tally() %>% 
  filter(n > 5) %>% 
  pull(celltype_tissue) %>% 
  unique()


for (genename in c(mast_wildrdw)) {
# for (genename in c(mast_wildrup)) {
  tissue <- "HYP"
  celltype <- "0568 VMH Nr5a1 Glut_3"

  gene_df <- get_gene_expression_df(
    seurat_obj,
    celltype_column = "supertype_name",
    celltype_value = celltype,
    gene = genename,
    tissue = tissue,
    return_format = "log2cpm") %>%
    mutate(sample = factor(sample, levels = unique(sample[order(microbiome, sample)]))) %>%
    glimpse()

  p_combined <- plot_gene_expression(
    gene_df,
    gene = genename,
    tissue = tissue,
    celltype_value = celltype
  )
  ggsave(
    glue("WildR_DOWN_{tissue}_{celltype}_{genename}_{Sys.Date()}.png"),
    p_combined,
    width = 8,
    height = 5
  )
}

```


Making a PCA for VMH Nr5a1 Glut_3 cells

```{r}
gene_df <- get_gene_expression_df_multi(seurat_obj, "supertype_name", "0568 VMH Nr5a1 Glut_3", 
                                    genes = c(mast_wildrup, mast_wildrdw))
# gene_df %>% glimpse()

gene_df %>% 
  ggplot(aes(x = Cntn6_log2cpm, y = Gabrg3_log2cpm)) +
  geom_point() +
  theme_bw()

```



```{r}
source(glue("{wkdir}/notebooks/R_scripts/_misc_functions.R"))

genename <- "Grm5"
tissue <- "HYP"
celltype <- "0568 VMH Nr5a1 Glut_3"

gene_df <- get_gene_expression_df(
  seurat_obj,
  celltype_column = "supertype_name",
  celltype_value = celltype,
  gene = genename,
  tissue = tissue,
  return_format = "log2cpm") %>%
  mutate(sample = factor(sample, levels = unique(sample[order(microbiome, sample)]))) %>%
  glimpse()
p_combined <- plot_gene_expression(
  gene_df,
  gene = genename,
  tissue = tissue,
  celltype_value = celltype
)
ggsave(
  glue("{genename}_{tissue}_{celltype}_{Sys.Date()}.png"),
  p_combined,
  width = 8,
  height = 5
)

# looking for pairwise correlations between top gene expressions of top hits
df2 <- get_gene_expression_df_multi(seurat_obj, "supertype_name", "0568 VMH Nr5a1 Glut_3", 
                                    genes = c(mast_wildrup, mast_wildrdw))
df2 %>% glimpse()

# Filter to log2cpm columns, compute pairwise correlation matrix
df2_log2cpm <- df2 %>% select(ends_with("_log2cpm"))
cor_mat <- cor(df2_log2cpm, use = "pairwise.complete.obs")
cor_df <- as.data.frame(cor_mat)
View(cor_df)

cor_df %>% 
  select(Cntn6_log2cpm)

# Long-format for heatmap
cor_long <- cor_mat %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene1") %>%
  pivot_longer(-gene1, names_to = "gene2", values_to = "correlation") %>%
  mutate(gene1 = str_remove(gene1, "_log2cpm"),
         gene2 = str_remove(gene2, "_log2cpm"))

# Hierarchical clustering to order axes (genes with similar correlation patterns grouped)
gene_order <- str_remove(colnames(cor_mat)[hclust(as.dist(1 - cor_mat))$order], "_log2cpm")
cor_long <- cor_long %>%
  mutate(gene1 = factor(gene1, levels = gene_order),
         gene2 = factor(gene2, levels = gene_order))

# Correlation heatmap
p_cor_heatmap <- ggplot(cor_long, aes(gene1, gene2, fill = correlation)) +
  geom_tile() +
  scale_fill_scico(palette = "vik", midpoint = 0, limits = c(-1, 1)) +
  labs(x = NULL, y = NULL, fill = "r") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_cor_heatmap

# using top hit genes to make a PCA






# for (genename in c(mast_wildrdw)) {
# # for (genename in c(mast_wildrup)) {
#   tissue <- "HYP"
#   celltype <- "0568 VMH Nr5a1 Glut_3"

#   gene_df <- get_gene_expression_df(
#     seurat_obj,
#     celltype_column = "supertype_name",
#     celltype_value = celltype,
#     gene = genename,
#     tissue = tissue,
#     return_format = "log2cpm") %>%
#     mutate(sample = factor(sample, levels = unique(sample[order(microbiome, sample)]))) %>%
#     glimpse()

#   p_combined <- plot_gene_expression(
#     gene_df,
#     gene = genename,
#     tissue = tissue,
#     celltype_value = celltype
#   )
#   ggsave(
#     glue("WildR_DOWN_{tissue}_{celltype}_{genename}_{Sys.Date()}.png"),
#     p_combined,
#     width = 8,
#     height = 5
#   )
# }


```


For each pair of samples calculate the KS test stat. 

```{r}
gene_df %>% glimpse()

genename <- "Cntn6"
tissue <- "HYP"
celltype <- "0568 VMH Nr5a1 Glut_3"
return_format <- "raw"

gene_df <- get_gene_expression_df(
  seurat_obj,
  celltype_column = "supertype_name",
  celltype_value = celltype,
  gene = genename,
  tissue = tissue,
  return_format = return_format) %>%
  mutate(sample = factor(sample, levels = unique(sample[order(microbiome, sample)]))) %>%
  glimpse()

gene_df %>% 
  ggplot(aes(x = Cntn6_log2cpm)) +
  # geom_histogram(alpha = 0.5) +
  geom_density(alpha = 0.5, aes(group = sample, fill = microbiome)) +
  facet_wrap(~microbiome) +
  theme_bw()


# Unique sample pairs: each comparison once, no self-comparisons (A vs B only, not B vs A)
samples <- as.character(unique(gene_df$sample))
sample_pairs <- combn(samples, 2) %>%
  t() %>%
  as.data.frame() %>%
  setNames(c("sample1", "sample2")) %>% 
  mutate(sample1_group = strex::str_before_first(sample1, "_"),
         sample2_group = strex::str_before_first(sample2, "_")) %>% 
  mutate(comparison = case_when(
    sample1_group == "WildR" & sample2_group == "WildR" ~ "WildR vs WildR",
    sample1_group == "SPF" & sample2_group == "SPF" ~ "SPF vs SPF",
    TRUE ~ "SPF vs WildR"
  )) %>% 
  glimpse()

# Use raw or log2cpm: pass expr_col = "Cntn6_raw" or "Cntn6_log2cpm", or use gene + return_format
ks_stats_df <- 
  compute_ks_stats(gene_df, sample_pairs, gene = genename, return_format = return_format) %>%
  glimpse()

# ANOVA + Tukey HSD post-hoc for pairwise comparisons
tukey_df <- ks_stats_df %>%
  rstatix::tukey_hsd(ks_stat ~ comparison) %>%
  rstatix::add_significance() %>%
  mutate(
    y.position = max(ks_stats_df$ks_stat) * (1 + seq_len(n()) * 0.12)
  )

p_ks <- ks_stats_df %>%
  ggplot(aes(x = comparison, y = ks_stat)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.85, width = 0.2) +
  ggpubr::stat_compare_means(method = "anova", label = "p.format") +
  ggpubr::stat_pvalue_manual(tukey_df, label = "p.adj.signif", hide.ns = FALSE) +
  labs(x = NULL, y = "KS Test Statistic", 
    title = glue("{genename} {tissue} {celltype} \nPairwise KS Test Statistics")) +
  theme_bw()

ggsave(
  glue("{genename}_{tissue}_{celltype}_ks_test_stats_{return_format}_{Sys.Date()}.png"),
  p_ks,
  width = 5,
  height = 5
)

```




Calculating the Odds ratio of each cell type for WildR vs SPF

```{r}
# (a/c) / (b/d)
# where a is the number of aged cells in the cluster, b is the number of adult cells in the cluster, c is the number of aged cells in the class minus a and d is the number of adult cells in the class minus b
# cell_type_name <- "class_name"

calculate_odds_ratio <- function(cell_type_name) {
  cell_counts_df <- seurat_obj@meta.data %>% 
    dplyr::select(cell_type_name = {{cell_type_name}}, microbiome, tissue) %>% 
    group_by(cell_type_name, microbiome, tissue) %>% 
    summarise(n_cells = n(), .groups = "drop")
    
  tissue_group_totals <- cell_counts_df %>% 
    group_by(tissue, microbiome) %>% 
    summarise(group_tissue_ncells = sum(n_cells), .groups = "drop") %>% 
    pivot_wider(names_from = microbiome, values_from = group_tissue_ncells, 
      names_prefix = "total_cells_") %>% 
    glimpse()

  cell_counts_df %>% 
    pivot_wider(names_from = microbiome, values_from = n_cells) %>% 
    left_join(tissue_group_totals) %>% 
    mutate(celltype_tissue = glue("{cell_type_name}_{tissue}")) %>% 
    mutate(odds_ratio = (WildR + 1 / (total_cells_WildR - WildR + 1)) / (SPF + 1 / (total_cells_SPF - SPF + 1))) %>% 
    glimpse()
}

# tst <- calculate_odds_ratio("class_name") %>% glimpse()
odds_ratio_df <- bind_rows(
  calculate_odds_ratio("class_name") %>% mutate(cell_type_level = "class"),
  calculate_odds_ratio("subclass_name") %>% mutate(cell_type_level = "subclass"),
  calculate_odds_ratio("supertype_name") %>% mutate(cell_type_level = "supertype")
  ) %>% 
  mutate(analysis = "Odds Ratio", log_odds_ratio = log2(odds_ratio)) %>% 
  arrange(desc(log_odds_ratio)) %>% 
  mutate(odds_ratio_hit = if_else(abs(log_odds_ratio) > 2.5, "Yes", "No")) %>% 
  glimpse()
# View(odds_ratio_df)

```


```{r}
# Running speckle/propeller testing for cell type proportion differences between the two 
# groups - this accounts for compositional data structure, variability in between samples, and prevents pseudoreplication

library(speckle)
library(limma)
metadata <- seurat_obj@meta.data %>% glimpse()

celltype_droplist <- readRDS(
  glue("{data_path}/seurat/snRNASeq/celltype_droplist_2025-12-30.rds")
  )

cell_prop_res <- list()
for (tiss in unique(metadata$tissue)) {
  metadata_tiss <- metadata %>% filter(tissue == tiss)
  tissue_droplist <- celltype_droplist %>% filter(tissue == tiss) %>% 
    left_join(allen_annot_long, by = c("celltype" = "cell_type_label"))
  
  for (cell_type_level in c("class_name", "subclass_name", "supertype_name")) {
  
  metadata_tiss_filtered <- metadata_tiss %>% 
    filter(.data[[cell_type_level]] %nin% unique(tissue_droplist$cell_type_name)) %>% 
    glimpse()
  
  cell_prop_res[[glue("{cell_type_level}_{tiss}")]] <- propeller(
    clusters = metadata_tiss[[cell_type_level]], 
    sample = metadata_tiss$sample,
    group = metadata_tiss$microbiome,
    transform = "logit"
    ) %>% 
    mutate(cell_type_level = cell_type_level, tissue = tiss)
    } %>% 
    arrange(P.Value) %>% 
    glimpse()
}

bind_rows(cell_prop_res) %>% 
  arrange(FDR) %>%
  glimpse()

# conclusion is there is no significant difference in cell type proportions between the two groups

```

NeuroeSTIMator results

```{r}
neuroest_beta_res_df <- readRDS(glue(
  "{wkdir}/data/interim/neuroestimator/",
  "neuroest_beta_lmm_res_df_snRNAseq_2026-01-11.rds"
)) %>% 
  mutate(analysis = "NeuroeSTIMator-GLMM", 
  neg_log10_p_value = -log10(p_value)
  )
neuroest_beta_res_df %>% glimpse()


```



## Q1: What cell types have the most DEGs, and are the most altered in state between WildR vs SPF?


```{r}
mast_df %>% glimpse()
edgeR_res %>% glimpse()
deseq2_res %>% glimpse()
augur_df %>% glimpse()

mast_deg_summary <- mast_df %>% 
    group_by(cell_type_name, tissue, cell_type_level) %>% 
    tally() %>% 
    mutate(analysis = "MAST")

edgeR_deg_summary <- edgeR_res %>% 
    group_by(cell_type_name, tissue, cell_type_level, celltype_tissue) %>% 
    tally() %>% 
    mutate(analysis = "EdgeR-Pseudobulk")

deseq2_deg_summary <- deseq2_res %>% 
    group_by(cell_type_name, tissue, cell_type_level, celltype_tissue) %>% 
    tally() %>% 
    mutate(analysis = "DESeq2-Pseudobulk")

```


```{r}
# Visualize a ranked list of cell types by number of DEGS in each tissue for each analysis
tiss <- "AMY"

deg_summary_plot <- bind_rows(mast_deg_summary, edgeR_deg_summary, deseq2_deg_summary) %>% 
  filter(tissue == tiss) %>% 
  filter(n > 1) %>% 
  mutate(analysis = factor(analysis, 
    levels = c("MAST", "EdgeR-Pseudobulk", "DESeq2-Pseudobulk"))) %>% 
  group_by(cell_type_level) %>% 
  arrange(n) %>% 
  mutate(cell_type_name = fct_reorder(cell_type_name, n)) %>% 
  glimpse()

# Get intersection of cell types present in both datasets
augur_tiss_celltypes <- augur_df %>% 
  filter(tissue == tiss) %>% 
  pull(cell_type_name) %>% 
  unique()

# Get the ordered cell types from DEG plot, preserving order within each level
deg_celltypes_ordered <- deg_summary_plot %>% 
  ungroup() %>% 
  filter(analysis == "MAST") %>% 
  select(cell_type_name, cell_type_level) %>% 
  distinct() %>% 
  mutate(cell_type_name = as.character(cell_type_name))

# Filter to only common cell types, preserving the order
common_celltypes_df <- deg_celltypes_ordered %>% 
  filter(cell_type_name %in% augur_tiss_celltypes)

# Create DEG summary plot (filtered to common cell types, preserving order within each level)
p_deg_summary <- deg_summary_plot %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ungroup() %>%
  ggplot(aes(x = n, y = cell_type_name)) +
    geom_bar(stat = "identity", fill = "gray50", width = 0.75) +
    # scale_x_log10() +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Number of DEGs (P-adjusted < 0.1)", y = "Cell Type") +
    theme(strip.text.y = element_blank(), 
      axis.text.x = element_text(angle = 45, hjust = 1))

# Create Augur lollipop chart (preserving order within each level)
p_augur_scores <- augur_df %>% 
  filter(tissue == tiss) %>% 
  mutate(analysis = "Augur") %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = auc, y = cell_type_name)) +
    geom_segment(aes(x = 0.5, xend = auc, y = cell_type_name, yend = cell_type_name), 
                 color = "gray50", linewidth = 0.5) +
    geom_point(aes(fill = auc), size = 3, shape = 21) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5) +
    scale_x_continuous(limits = c(0.5, 0.8), expand = c(0, 0)) +
    theme_bw() +
    facet_grid(cell_type_level ~  analysis, scales = "free", space = "free_y") +
    labs(x = "AUROC", y = "", color = "AUROC") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

log_odds_ratio_plot <- odds_ratio_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_ratio, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = odds_ratio_hit), width = 0.75) +
    scale_fill_manual(values = c("Yes" = "darkred", "No" = "gray50")) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log2 Odds Ratio (WildR / SPF)", y = NULL) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

neuroest_beta_plot <- neuroest_beta_res_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_effect, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = neg_log10_p_value), 
      width = 0.75) +
    scale_fill_viridis_c(option = "cividis", direction = -1) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log Odds Effect (WildR / SPF)", y = NULL, fill = "-log10(P-value)") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank()
          )

# Align plots using patchwork (which should align by shared y-axis)
p_combined <- p_deg_summary + p_augur_scores + log_odds_ratio_plot + neuroest_beta_plot +
  plot_layout(widths = c(3, 1, 1, 1), guides = "collect") +
  plot_annotation(title = "Differential Expression Analysis",
                  subtitle = tiss)
# p_combined

# ggsave(
#   glue("{fig_path}/DEG/snRNASeq/mast_deg_summary_combined_{tiss}_{Sys.Date()}.png"), 
#   p_combined, width = 18, height = 16, dpi = 300)

```

Summary for HYPOTHALAMUS

```{r}
tiss <- "HYP"

deg_summary_plot <- bind_rows(mast_deg_summary, edgeR_deg_summary, deseq2_deg_summary) %>% 
  filter(tissue == tiss) %>% 
  filter(n > 1) %>% 
  mutate(analysis = factor(analysis, 
    levels = c("MAST", "EdgeR-Pseudobulk", "DESeq2-Pseudobulk"))) %>% 
  group_by(cell_type_level) %>% 
  arrange(n) %>% 
  left_join(allen_annot_long) %>% 
  # mutate(cell_type_name = fct_reoder(cell_type_name, cell_type_name)) %>% 
  glimpse()

# Get intersection of cell types present in both datasets
augur_tiss_celltypes <- augur_df %>% 
  filter(tissue == tiss) %>% 
  pull(cell_type_name) %>% 
  unique()

# Get the ordered cell types from DEG plot, preserving order within each level
deg_celltypes_ordered <- deg_summary_plot %>% 
  ungroup() %>% 
  filter(analysis == "MAST") %>% 
  select(cell_type_name, cell_type_level) %>% 
  distinct() %>% 
  mutate(cell_type_name = as.character(cell_type_name))

# Filter to only common cell types, preserving the order
common_celltypes_df <- deg_celltypes_ordered %>% 
  filter(cell_type_name %in% augur_tiss_celltypes)

# Create DEG summary plot (filtered to common cell types, preserving order within each level)
p_deg_summary <- deg_summary_plot %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ungroup() %>%
  ggplot(aes(x = n, y = cell_type_name)) +
    geom_bar(stat = "identity", fill = "gray50", width = 0.75) +
    # scale_x_log10() +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Number of DEGs (P-adjusted < 0.1)", y = "Cell Type") +
    theme(strip.text.y = element_blank(), 
      axis.text.x = element_text(angle = 45, hjust = 1))

# Create Augur lollipop chart (preserving order within each level)
p_augur_scores <- augur_df %>% 
  filter(tissue == tiss) %>% 
  mutate(analysis = "Augur") %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = auc, y = cell_type_name)) +
    geom_segment(aes(x = 0.5, xend = auc, y = cell_type_name, yend = cell_type_name), 
                 color = "gray50", linewidth = 0.5) +
    geom_point(aes(fill = auc), size = 3, shape = 21) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5) +
    scale_x_continuous(limits = c(0.5, 0.8), expand = c(0, 0)) +
    theme_bw() +
    facet_grid(cell_type_level ~  analysis, scales = "free_y", space = "free") +
    labs(x = "AUROC", y = "", color = "AUROC") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

log_odds_ratio_plot <- odds_ratio_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_ratio, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = odds_ratio_hit), width = 0.75) +
    scale_fill_manual(values = c("Yes" = "darkred", "No" = "gray50")) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log2 Odds Ratio (WildR / SPF)", y = NULL) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

neuroest_beta_plot <- neuroest_beta_res_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_effect, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = neg_log10_p_value), 
      width = 0.75) +
    scale_fill_viridis_c(option = "cividis", direction = -1) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log Odds Effect (WildR / SPF)", y = NULL, fill = "-log10(P-value)") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank()
          )

# Align plots using patchwork (which should align by shared y-axis)
p_combined <- p_deg_summary + p_augur_scores + log_odds_ratio_plot + neuroest_beta_plot +
  plot_layout(widths = c(3, 1, 1, 1), guides = "collect") +
  plot_annotation(title = "Differential Expression Analysis",
                  subtitle = tiss)

# p_combined
# ggsave(
#   glue("{fig_path}/DEG/snRNASeq/mast_deg_summary_combined_{tiss}_{Sys.Date()}.png"), 
#   p_combined, width = 18, height = 19, dpi = 300)


```



REMAKING SUMMARY FIGURES WITH BOTH Pvalue and FDR thresholds

```{r}
mast_df_pvalue <- bind_rows(class_mast_df, subclass_mast_df, supertype_mast_df) %>% 
  filter(pr_chisq < 0.05) %>% 
    glimpse()

edgeR_res_pvalue <- readRDS(
  glue("{data_path}/edgeR/snRNASeq/edgeR_signif_hits_lrt_res_2026-01-03.rds")) %>% 
  mutate(
    tissue = str_after_last(celltype_tissue, "_"), 
    cell_type = str_before_last(celltype_tissue, "_"), 
    cell_type_label = cell_type) %>% 
  filter(PValue < 0.05) %>% 
  left_join(allen_annot_long) %>% 
  glimpse()

# Deseq2 Results
deseq2_res_pvalue <- readRDS(
  glue("{data_path}/DESeq2/snRNASeq/deseq2_signif_hits_res_2026-01-03.rds")) %>% 
  mutate(
    tissue = str_after_last(celltype_tissue, "_"), 
    cell_type = str_before_last(celltype_tissue, "_"), 
    cell_type_label = cell_type) %>% 
  left_join(allen_annot_long) %>% 
  filter(pvalue < 0.05) %>% 
  left_join(allen_annot_long) %>% 
  glimpse()

mast_deg_summary_pvalue <- mast_df_pvalue %>% 
    group_by(cell_type_name, tissue, cell_type_level) %>% 
    tally() %>% 
    mutate(analysis = "MAST")
edgeR_deg_summary_pvalue <- edgeR_res_pvalue %>% 
    group_by(cell_type_name, tissue, cell_type_level, celltype_tissue) %>% 
    tally() %>% 
    mutate(analysis = "EdgeR-Pseudobulk")
deseq2_deg_summary_pvalue <- deseq2_res_pvalue %>% 
    group_by(cell_type_name, tissue, cell_type_level, celltype_tissue) %>% 
    tally() %>% 
    mutate(analysis = "DESeq2-Pseudobulk")

```

  

```{r}
# Visualize a ranked list of cell types by number of DEGS in each tissue for each analysis
tiss <- "AMY"

deg_summary_plot <- bind_rows(mast_deg_summary, edgeR_deg_summary, deseq2_deg_summary) %>% 
  filter(tissue == tiss) %>% 
  filter(n > 1) %>% 
  mutate(analysis = factor(analysis, 
    levels = c("MAST", "EdgeR-Pseudobulk", "DESeq2-Pseudobulk"))) %>% 
  group_by(cell_type_level) %>% 
  arrange(n) %>% 
  mutate(pthres = "FDR < 0.1") %>% 
  glimpse()

deg_summary_plot_pval <- bind_rows(
    mast_deg_summary_pvalue, edgeR_deg_summary_pvalue, deseq2_deg_summary_pvalue) %>% 
    left_join(allen_annot_long) %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% deg_summary_plot$cell_type_name) %>% 
  mutate(analysis = factor(analysis, 
    levels = c("MAST", "EdgeR-Pseudobulk", "DESeq2-Pseudobulk"))) %>% 
  mutate(pthres = "P-value < 0.05") %>% 
  glimpse()

# Get intersection of cell types present in both datasets
augur_tiss_celltypes <- augur_df %>% 
  filter(tissue == tiss) %>% 
  pull(cell_type_name) %>% 
  unique()

# Get the ordered cell types from DEG plot, preserving order within each level
deg_celltypes_ordered <- deg_summary_plot %>% 
  ungroup() %>% 
  filter(analysis == "MAST") %>% 
  select(cell_type_name, cell_type_level) %>% 
  distinct() %>% 
  mutate(cell_type_name = as.character(cell_type_name))

# Filter to only common cell types, preserving the order
common_celltypes_df <- deg_celltypes_ordered %>% 
  filter(cell_type_name %in% augur_tiss_celltypes)

# Create DEG summary plot (filtered to common cell types, preserving order within each level)
p_deg_summary <- bind_rows(deg_summary_plot_pval, deg_summary_plot) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  # mutate(pthres = factor(pthres, levels = c("FDR < 0.1", "P-value < 0.05"))) %>% 
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ungroup() %>%
  ggplot(aes(x = n, y = cell_type_name, fill = pthres)) +
    geom_bar(stat = "identity", position = "identity", width = 0.8) +
    scale_fill_manual(values = c("P-value < 0.05" = "darkgrey", "FDR < 0.1" = "darkred")) +
    scale_x_log10() +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Number of DEGs", y = "Cell Type", fill = NULL) +
    theme(strip.text.y = element_blank(), 
      axis.text.x = element_text(angle = 45, hjust = 1))

# Create Augur lollipop chart (preserving order within each level)
p_augur_scores <- augur_df %>% 
  filter(tissue == tiss) %>% 
  mutate(analysis = "Augur") %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = auc, y = cell_type_name)) +
    geom_segment(aes(x = 0.5, xend = auc, y = cell_type_name, yend = cell_type_name), 
                 color = "gray50", linewidth = 0.5) +
    geom_point(aes(fill = auc), size = 3, shape = 21) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5) +
    scale_x_continuous(limits = c(0.5, 0.8), expand = c(0, 0)) +
    theme_bw() +
    facet_grid(cell_type_level ~  analysis, scales = "free", space = "free_y") +
    labs(x = "AUROC", y = "", color = "AUROC") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

log_odds_ratio_plot <- odds_ratio_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_ratio, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = odds_ratio_hit), width = 0.75) +
    scale_fill_manual(values = c("Yes" = "darkred", "No" = "gray50")) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log2 Odds Ratio (WildR / SPF)", y = NULL) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

neuroest_beta_plot <- neuroest_beta_res_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_effect, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = neg_log10_p_value), 
      width = 0.75) +
    scale_fill_viridis_c(option = "cividis", direction = -1) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log Odds Effect (WildR / SPF)", y = NULL, fill = "-log10(P-value)") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank()
          )

# Align plots using patchwork (which should align by shared y-axis)
p_combined <- p_deg_summary + p_augur_scores + log_odds_ratio_plot + neuroest_beta_plot +
  plot_layout(widths = c(3, 1, 1, 1), guides = "collect") +
  plot_annotation(title = "Differential Expression Analysis",
                  subtitle = tiss)
# p_combined

ggsave(
  glue("{fig_path}/DEG/snRNASeq/mast_deg_summary_combined_{tiss}_pvalue_fdr_{Sys.Date()}.png"), 
  p_combined, width = 18, height = 16, dpi = 300)

```

Summary for HYPOTHALAMUS

```{r}
tiss <- "HYP"

deg_summary_plot <- bind_rows(mast_deg_summary, edgeR_deg_summary, deseq2_deg_summary) %>% 
  filter(tissue == tiss) %>% 
  filter(n > 1) %>% 
  mutate(analysis = factor(analysis, 
    levels = c("MAST", "EdgeR-Pseudobulk", "DESeq2-Pseudobulk"))) %>% 
  group_by(cell_type_level) %>% 
  arrange(n) %>% 
  mutate(pthres = "FDR < 0.1") %>% 
  glimpse()

deg_summary_plot_pval <- bind_rows(
    mast_deg_summary_pvalue, edgeR_deg_summary_pvalue, deseq2_deg_summary_pvalue) %>% 
    left_join(allen_annot_long) %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% deg_summary_plot$cell_type_name) %>% 
  mutate(analysis = factor(analysis, 
    levels = c("MAST", "EdgeR-Pseudobulk", "DESeq2-Pseudobulk"))) %>% 
  mutate(pthres = "P-value < 0.05") %>% 
  glimpse()

# Get intersection of cell types present in both datasets
augur_tiss_celltypes <- augur_df %>% 
  filter(tissue == tiss) %>% 
  pull(cell_type_name) %>% 
  unique()

# Get the ordered cell types from DEG plot, preserving order within each level
deg_celltypes_ordered <- deg_summary_plot %>% 
  ungroup() %>% 
  filter(analysis == "MAST") %>% 
  select(cell_type_name, cell_type_level) %>% 
  distinct() %>% 
  mutate(cell_type_name = as.character(cell_type_name))

# Filter to only common cell types, preserving the order
common_celltypes_df <- deg_celltypes_ordered %>% 
  filter(cell_type_name %in% augur_tiss_celltypes)

# Create DEG summary plot (filtered to common cell types, preserving order within each level)
p_deg_summary <- bind_rows(deg_summary_plot_pval, deg_summary_plot) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  # mutate(pthres = factor(pthres, levels = c("FDR < 0.1", "P-value < 0.05"))) %>% 
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ungroup() %>%
  ggplot(aes(x = n, y = cell_type_name, fill = pthres)) +
    geom_bar(stat = "identity", position = "identity", width = 0.7) +
    scale_fill_manual(values = c("P-value < 0.05" = "darkgrey", "FDR < 0.1" = "darkred")) +
    scale_x_log10() +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Number of DEGs", y = "Cell Type", fill = NULL) +
    theme(strip.text.y = element_blank(), 
      axis.text.x = element_text(angle = 45, hjust = 1))

# Create Augur lollipop chart (preserving order within each level)
p_augur_scores <- augur_df %>% 
  filter(tissue == tiss) %>% 
  mutate(analysis = "Augur") %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = auc, y = cell_type_name)) +
    geom_segment(aes(x = 0.5, xend = auc, y = cell_type_name, yend = cell_type_name), 
                 color = "gray50", linewidth = 0.5) +
    geom_point(aes(fill = auc), size = 3, shape = 21) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5) +
    scale_x_continuous(limits = c(0.5, 0.8), expand = c(0, 0)) +
    theme_bw() +
    facet_grid(cell_type_level ~  analysis, scales = "free", space = "free_y") +
    labs(x = "AUROC", y = "", color = "AUROC") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

log_odds_ratio_plot <- odds_ratio_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_ratio, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = odds_ratio_hit), width = 0.75) +
    scale_fill_manual(values = c("Yes" = "darkred", "No" = "gray50")) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log2 Odds Ratio (WildR / SPF)", y = NULL) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

neuroest_beta_plot <- neuroest_beta_res_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_effect, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = neg_log10_p_value), 
      width = 0.75) +
    scale_fill_viridis_c(option = "cividis", direction = -1) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log Odds Effect (WildR / SPF)", y = NULL, fill = "-log10(P-value)") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank()
          )

# Align plots using patchwork (which should align by shared y-axis)
p_combined <- p_deg_summary + p_augur_scores + log_odds_ratio_plot + neuroest_beta_plot +
  plot_layout(widths = c(3, 1, 1, 1), guides = "collect") +
  plot_annotation(title = "Differential Expression Analysis",
                  subtitle = tiss)
# p_combined

ggsave(
  glue("{fig_path}/DEG/snRNASeq/mast_deg_summary_combined_{tiss}_pvalue_fdr_{Sys.Date()}.png"), 
  p_combined, width = 18, height = 19, dpi = 300)


```


## Pulling out cell types of interest from this panel

```{r}
# tiss <- "HYP"
# cell_type_regex <- "PVpo-VMPO-MPN|AVPV-MEPO-SFO|LHA-MEA Otp"
tiss <- "AMY"
cell_type_regex <- "MEA-BST Sox6|MEA-BST Lhx6|COAa-PAA-MEA"

deg_summary_plot <- bind_rows(mast_deg_summary, edgeR_deg_summary, deseq2_deg_summary) %>% 
  filter(tissue == tiss) %>% 
  filter(n > 1) %>% 
  mutate(analysis = factor(analysis, 
    levels = c("MAST", "EdgeR-Pseudobulk", "DESeq2-Pseudobulk"))) %>% 
  group_by(cell_type_level) %>% 
  arrange(n) %>% 
  mutate(pthres = "FDR < 0.1") %>% 
  glimpse()

deg_summary_plot_pval <- bind_rows(
    mast_deg_summary_pvalue, edgeR_deg_summary_pvalue, deseq2_deg_summary_pvalue) %>% 
    left_join(allen_annot_long) %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% deg_summary_plot$cell_type_name) %>% 
  mutate(analysis = factor(analysis, 
    levels = c("MAST", "EdgeR-Pseudobulk", "DESeq2-Pseudobulk"))) %>% 
  mutate(pthres = "P-value < 0.05") %>% 
  glimpse()

# Get intersection of cell types present in both datasets
augur_tiss_celltypes <- augur_df %>% 
  filter(tissue == tiss) %>% 
  pull(cell_type_name) %>% 
  unique()

# Get the ordered cell types from DEG plot, preserving order within each level
deg_celltypes_ordered <- deg_summary_plot %>% 
  ungroup() %>% 
  filter(analysis == "MAST") %>% 
  select(cell_type_name, cell_type_level) %>% 
  distinct() %>% 
  mutate(cell_type_name = as.character(cell_type_name)) %>% 
  filter(grepl(cell_type_regex, cell_type_name)) %>% 
  glimpse()

# Filter to only common cell types, preserving the order
common_celltypes_df <- deg_celltypes_ordered %>% 
  filter(cell_type_name %in% augur_tiss_celltypes)

# Create DEG summary plot (filtered to common cell types, preserving order within each level)
p_deg_summary <- bind_rows(deg_summary_plot_pval, deg_summary_plot) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  # mutate(pthres = factor(pthres, levels = c("FDR < 0.1", "P-value < 0.05"))) %>% 
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ungroup() %>%
  ggplot(aes(x = n, y = cell_type_name, fill = pthres)) +
    geom_bar(stat = "identity", position = "identity", width = 0.7) +
    scale_fill_manual(values = c("P-value < 0.05" = "darkgrey", "FDR < 0.1" = "darkred")) +
    scale_x_log10() +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Number of DEGs", y = "Cell Type", fill = NULL) +
    theme(strip.text.y = element_blank(), 
      axis.text.x = element_text(angle = 45, hjust = 1))

# Create Augur lollipop chart (preserving order within each level)
p_augur_scores <- augur_df %>% 
  filter(tissue == tiss) %>% 
  mutate(analysis = "Augur") %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = auc, y = cell_type_name)) +
    geom_segment(aes(x = 0.5, xend = auc, y = cell_type_name, yend = cell_type_name), 
                 color = "gray50", linewidth = 0.5) +
    geom_point(aes(fill = auc), size = 3, shape = 21) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0.5) +
    scale_x_continuous(limits = c(0.5, 0.8), expand = c(0, 0)) +
    theme_bw() +
    facet_grid(cell_type_level ~  analysis, scales = "free", space = "free_y") +
    labs(x = "AUROC", y = "", color = "AUROC") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

log_odds_ratio_plot <- odds_ratio_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_ratio, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = odds_ratio_hit), width = 0.75) +
    scale_fill_manual(values = c("Yes" = "darkred", "No" = "gray50")) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log2 Odds Ratio (WildR / SPF)", y = NULL) +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.position = "none"
          )

neuroest_beta_plot <- neuroest_beta_res_df %>% 
  filter(tissue == tiss) %>% 
  filter(cell_type_name %in% common_celltypes_df$cell_type_name) %>%
  mutate(cell_type_name = factor(cell_type_name, 
                                 levels = deg_celltypes_ordered$cell_type_name)) %>%
  ggplot(aes(x = log_odds_effect, y = cell_type_name)) +
    geom_bar(stat = "identity", aes(fill = neg_log10_p_value), 
      width = 0.75) +
    scale_fill_viridis_c(option = "cividis", direction = -1) +
    theme_bw() +
    facet_grid(cell_type_level ~ analysis, scales = "free", space = "free_y") +
    labs(x = "Log Odds Effect (WildR / SPF)", y = NULL, fill = "-log10(P-value)") +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank()
          )

# Align plots using patchwork (which should align by shared y-axis)
p_combined <- p_deg_summary + p_augur_scores + log_odds_ratio_plot + neuroest_beta_plot +
  plot_layout(widths = c(3, 1, 1, 1), guides = "collect") +
  plot_annotation(title = "Differential Expression Analysis",
                  subtitle = tiss)
# p_combined

ggsave(
  glue("{fig_path}/DEG/snRNASeq/",
  "mast_deg_summary_combined_{tiss}_pvalue_fdr_{cell_type_regex}_{Sys.Date()}.png"),
  p_combined, width = 18, height = 4, dpi = 300)


```




Spot checking that tissue labels are correct for the EdgeR data - it somehow seems that EdgeR trends in HYP match whats going on in MAST(AMY) data better than in MAST (HYP) data


```{r}
edger_dir <- glue("{data_path}/edgeR/snRNASeq")
mast_resdir <- glue("{data_path}/mast/snRNASeq")
# View(allen_annot_df)
# CS20230722_CLAS_12_AMY

mast_12_hyp <- readRDS(glue("{mast_resdir}/class_label/CS20230722_CLAS_12_HYP_2025-12-31.rds")) %>% 
  dplyr::select(gene_id = primerid, PValue = `Pr(>Chisq)`, FDR = p_adjust, coef) %>% 
    dplyr::rename_with(~ paste0(.x, "_HYP_MAST"), .cols = c(PValue, FDR, coef)) %>% 
  glimpse()

pb_12_hyp <- readRDS(glue("{edger_dir}/class_label/CS20230722_CLAS_12_HYP/res_lrt_2025-12-30.rds")) %>% 
  select(logFC, PValue, FDR) %>% 
  dplyr::rename_all(~ paste0(.x, "_HYP")) %>% 
  rownames_to_column("gene_id") %>% 
  glimpse()
pb_12_amy <- readRDS(glue("{edger_dir}/class_label/CS20230722_CLAS_12_AMY/res_lrt_2025-12-30.rds")) %>% 
  select(logFC, PValue, FDR) %>% 
  dplyr::rename_all(~ paste0(.x, "_AMY")) %>% 
  rownames_to_column("gene_id") %>% 
  glimpse()


merged_12_data <- full_join(pb_12_hyp, pb_12_amy, by = "gene_id") %>% 
  full_join(mast_12_hyp) %>% 
  glimpse()

# I want to know if the labels on the edgeR results are correct for tissue, looking to see if the results with MAST data are more consistent in the right tissue (HYP) vs AMY
merged_12_data %>% 
  ggplot(aes(x = logFC_HYP, y = coef_HYP_MAST)) +
    geom_point() +
    theme_bw() +
    labs(x = "EdgeR LogFC", y = "MAST LogFC")
  
merged_12_data %>% 
  ggplot(aes(x = logFC_AMY, y = coef_HYP_MAST)) +
    geom_point() +
    theme_bw() +
    labs(x = "EdgeR LogFC", y = "MAST LogFC")


```



Plotting counts of significant hits across tissue groups

```{r}

mast_deg_summary_wide <- mast_deg_summary %>% dplyr::rename(n_mast = n) %>% select(-analysis) 
edgeR_deg_summary_wide <- edgeR_deg_summary %>% dplyr::rename(n_edgeR = n) %>% select(-analysis) 

mast_edger_summary <- full_join(
  mast_deg_summary_wide, 
  edgeR_deg_summary_wide) %>% 
  glimpse()

mast_edger_summary %>% 
  ggplot(aes(x = n_mast, y = n_edgeR)) +
    geom_point() +
    theme_bw() +
    scale_x_log10() +
    scale_y_log10() +
    geom_smooth(method = "lm") +
    facet_wrap(~ tissue) +
    labs(x = "Number of DEGs (MAST)", y = "Number of DEGs (EdgeR)")

# swapping tissue labels to test if its a better fit 
edgeR_deg_summary_wide_swapped <- edgeR_deg_summary %>% dplyr::rename(n_edgeR = n) %>% select(-analysis)  %>% 
  mutate(tissue = case_when(
    tissue == "AMY" ~ "HYP",
    tissue == "HYP" ~ "AMY",
    TRUE ~ tissue
  ))

mast_edger_summary_swapped <- full_join(
  mast_deg_summary_wide, 
  edgeR_deg_summary_wide_swapped) %>% 
  glimpse()

mast_edger_summary_swapped %>% 
  ggplot(aes(x = n_mast, y = n_edgeR)) +
    geom_point() +
    theme_bw() +
    scale_x_log10() +
    scale_y_log10() +
    geom_smooth(method = "lm") +
    facet_wrap(~ tissue) +
    labs(x = "Number of DEGs (MAST)", y = "Number of DEGs (EdgeR)")



```



### Q1.A Is there any trend between how many cells are used in MAST and the number of DEGs? 

Some cell types appear to have a bunch of differences, is this just because of greater cell counts going into the analysis?

```{r}
library(ggpubr)
library(ggrepel)

get_mast_cell_counts <- function(cell_type_column, celltype, tissue)  {
    get_time <- function() print(format(Sys.time(), "%Y-%m-%d_%H:%M:%S"))
    # Filter CELLS using metadata directly
    message(glue("{get_time()} Filtering cells"))
    cells_to_keep <- seurat_obj@meta.data[[cell_type_column]] == celltype & 
                    seurat_obj@meta.data$tissue == tissue
    return(sum(cells_to_keep))
}

celltype_droplist <- readRDS(
  glue("{data_path}/seurat/snRNASeq/celltype_droplist_2025-12-30.rds")
  )

metadata_df <- seurat_obj@meta.data %>% glimpse()
class_batchdata <- 
    dplyr::select(metadata_df, celltype = class_label, tissue) %>% 
    distinct() %>%
    mutate(cell_type_column = "class_label")
subclass_batchdata <- 
    dplyr::select(metadata_df, celltype = subclass_label, tissue) %>% 
    distinct() %>%
    mutate(cell_type_column = "subclass_label")
supertype_batchdata <- 
    dplyr::select(metadata_df, celltype = supertype_label, tissue) %>% 
    distinct() %>%
    mutate(cell_type_column = "supertype_label")

batchparams_df <- 
    bind_rows(class_batchdata, subclass_batchdata, supertype_batchdata) %>% 
    left_join(celltype_droplist) %>%
    filter(is.na(dropme)) %>% 
    dplyr::select(cell_type_column, celltype, tissue) %>% 
    glimpse()

mast_cell_counts_df <- batchparams_df %>%
  mutate(n_cells = pmap_dbl(list(cell_type_column, celltype, tissue), get_mast_cell_counts)) %>%
  mutate(cell_type_label = celltype) %>% 
  left_join(allen_annot_long) %>% 
  glimpse()

# new update subsamples down to 2000 cells per group max
mast_cell_counts_df %<>%
  mutate(n_cells = if_else(n_cells > 4000, 4000, n_cells)) %>% 
  glimpse()

mast_deg_summary_strict <- bind_rows(class_mast_df, subclass_mast_df, supertype_mast_df) %>%
    # filter(pr_chisq < 0.05) %>% 
    filter(p_adjust < 0.01) %>% 
    group_by(cell_type_name, tissue, cell_type_level, celltype_tissue) %>% 
    tally() %>% 
    mutate(analysis = "MAST") %>% 
    glimpse()

mast_deg_summary_wcounts <- mast_deg_summary_strict %>% 
  left_join(mast_cell_counts_df) %>% 
  glimpse()

p_mast_deg_summary_wcounts <- mast_deg_summary_wcounts %>% 
  ggplot(aes(x = n_cells, y = n)) +
    geom_point() +
    geom_smooth(method = "lm") +
    stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.97) +
    stat_regline_equation(label.x.npc = 0.05, label.y.npc = 0.91) +
    theme_bw() +
    # geom_vline(xintercept = 500, linetype = "dashed") +
    # scale_x_log10() +
    # scale_y_log10() +
    # facet_grid(cell_type_level ~ tissue) +
    facet_wrap( ~ tissue , scales = "free") +
    labs(x = "Number of Cells", y = "Number of DEGs")

ggsave(
  glue("{fig_path}/DEG/snRNASeq/mast_deg_summary_wcounts_{Sys.Date()}.png"), 
  p_mast_deg_summary_wcounts, width = 10, height = 6, dpi = 300)


# Identify top 5 cell types with most DEGs per tissue
top5_celltypes <- mast_deg_summary_wcounts %>%
  group_by(tissue) %>%
  slice_max(n, n = 10) %>%
  ungroup()

p_mast_deg_summary_wcounts_labeled <- mast_deg_summary_wcounts %>% 
  ggplot(aes(x = n_cells, y = n)) +
    geom_point() +
    geom_smooth(method = "lm") +
    stat_cor(method = "pearson", label.x.npc = 0.05, label.y.npc = 0.97) +
    stat_regline_equation(label.x.npc = 0.05, label.y.npc = 0.91) +
    geom_label_repel(
      data = top5_celltypes,
      aes(label = cell_type_name),
      alpha = 0.8,
      force = 10,
      force_pull = 0,
      nudge_x = 0.25,
      nudge_y = 0.25,
      segment.curvature = 0.3,
      segment.ncp = 3,
      segment.angle = 10,
      min.segment.length = 0,
      box.padding = 0.5,
      point.padding = 0.3,
      max.overlaps = Inf
    ) +
    theme_bw() +
    facet_wrap( ~ tissue , scales = "free") +
    labs(x = "Number of Cells", y = "Number of DEGs")

ggsave(
  glue("{fig_path}/DEG/snRNASeq/mast_deg_summary_wcounts_labeled_{Sys.Date()}.png"), 
  p_mast_deg_summary_wcounts_labeled, width = 10, height = 6, dpi = 300)

# s
# looks like theres a correlation in the HYP but not the AMY for number of cells in analysis and number of DEGs returned for that cell type with MAST.

```





```{r}

```

