---
title: "Metabolic Flux Analysis"
editor: source
author: "Joe Boktor"
date: '2026-01-24'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

```{r}
#| label: setup-display
#| include: false

# Normal render/preview: FALSE — no Seurat load, no matrix export, no sbatch.
# Set TRUE only when intentionally re-running the full scFEA pipeline.
run_analysis <- FALSE
```

```{r}
#| label: setup
#| eval: !expr run_analysis

library(Seurat)
library(Matrix)
library(irlba)
library(scico)
library(tidyverse)
library(magrittr)
library(patchwork)
library(glue)
library(presto)
library(hdf5r)
library(tidyseurat)
library(Augur)
library(batchtools)
library(tictoc)
library(future)
library(strex)
library(cowplot)
library(aplot)
library(SingleCellExperiment)
library(BiocParallel)
library(ggside)
theme_set(theme_bw())

home_dir <- "/resnick/groups/mthomson/jboktor"

wkdir <- glue("{home_dir}/WILDRxSPF_brains")
data_path <- glue("{wkdir}/data/interim")
fig_path <- glue("{wkdir}/figures")
abca_color_pal <- readRDS(glue("{wkdir}/data/interim/abca_color_pal.rds"))
scFEA_gitpath <- glue("{home_dir}/git/scFEA")
scFEA_input_path <- glue("{wkdir}/data/input/scFEA")
scFEA_output_path <- glue("{wkdir}/data/interim/scFEA/snRNASeq")
# dir.create(scFEA_output_path, showWarnings = FALSE, recursive = TRUE)

source(glue("{wkdir}/notebooks/R_scripts/_misc_functions.R"))

# Set memory limit for future
options(future.globals.maxSize = 200000 * 1024^2)  # Set to ~200GB
# Use more memory-efficient serialization
options(future.serialization.progressr = FALSE)
options(future.serialization.progressr.enable = FALSE)

# future::plan(multisession, workers = 3)
# future::plan(sequential)
```


## Analysis Prep

Loading Seurat object

```{r}
#| eval: !expr run_analysis

seurat_obj <- readRDS( 
    glue("{data_path}/seurat/snRNASeq/",
    "seurat_obj_onlyparsefilters_celltyped_2025-12-27.rds")
    )
seurat_obj
# Subset the seurat object to only keep cells that pass the QC filters
seurat_obj <- seurat_obj %>% subset(subset = keep_cell)
seurat_obj@meta.data %>% glimpse()

allen_annot_df <- seurat_obj@meta.data %>% 
  select(class_name, class_label, subclass_name, 
    subclass_label, supertype_name, supertype_label) %>% 
  distinct() %>% 
  glimpse()

allen_annot_long <- allen_annot_df %>% 
  pivot_longer(
    cols = everything(),
    names_to = c("cell_type_level", ".value"),
    names_pattern = "(class|subclass|supertype)_(name|label)"
  ) %>% 
  dplyr::rename(cell_type_name = name, cell_type_label = label) %>% 
  distinct() %>% 
  glimpse()

```


Git repo for scFEA is cloned into this path `scFEA_gitpath`.


Save a cellxgene matrix for each mouse sample separately, For each sample matrix:

1. CPM normalize, log2 transform
2. Filter out specific genes in metabolic modules
3. Save matrix in our data path



```{r}
#| eval: !expr run_analysis

# Load Metabolic Gene List (Clean names, e.g., "Hk2", "G6pd")
metabolic_genes <- read.delim(glue("{scFEA_input_path}/scFEA.mouse.genes.txt"), header = FALSE) %>% 
  pull(V1)

# generate_cpm_matrix <- function(seurat_obj, tissue) {
tissues <- c("AMY", "HYP")
# ts <- "AMY"
for (ts in tissues) {
  # STEP 1: Get Raw Data & Calculate Library Size
  seurat_ts <- seurat_obj %>% subset(subset = tissue == ts)
  # get a list of celltypes for this tissue
  celltype_counts <- seurat_ts@meta.data$class_label %>% table()
  celltype_labels <- names(celltype_counts)[celltype_counts > 100]
  
  for (celltype in celltype_labels) {
    output_file <- glue("{scFEA_input_path}/snRNASeq/snRNASeq_log2cpm_{ts}_{celltype}.csv")
    if (file.exists(output_file)) {
      message(glue("Skipping {celltype} in {ts} because it already exists"))
      next
    }
    message(glue("Processing {celltype} in {ts}"))
    seurat_celltype <- seurat_ts %>% subset(subset = class_label == celltype)
    # check if the celltype has at least 100 cells
    if (ncol(seurat_celltype) < 100) {
      message(glue("Number of cells in {celltype} in {ts}: {ncol(seurat_celltype)}"))
      message(glue("Skipping {celltype} in {ts} because it has less than 100 cells"))
      next
    }
    counts <- Seurat::GetAssayData(seurat_celltype, layer = "counts")
    total_counts <- Matrix::colSums(counts)

    # STEP 2: Filter for Metabolic Genes
    metabolism_gene_lgl <- seurat_celltype[["RNA"]][[]]$gene_name %in% metabolic_genes
    counts_filtered <- counts[metabolism_gene_lgl, ]
    # Sanity Check
    message(glue("Original Genes: {nrow(counts)}"))
    message(glue("Filtered Metabolic Genes: {nrow(counts_filtered)}"))
    message(glue("Genes Removed: {sum(!metabolism_gene_lgl)}"))


    # STEP 3: Aggregate Isoforms (The Matrix Multiplication Method)
    # 3a. Clean Gene Names (Remove suffixes like -GRCm39)
    clean_genes <- strex::str_before_first(rownames(counts_filtered), "-GRCm39")

    # 3b. create the Transformation Matrix
    # as.factor sorts levels alphabetically -> Rows of agg_mat will be sorted!
    clean_genes_factor <- as.factor(clean_genes)
    agg_mat <- as(clean_genes_factor, "sparseMatrix") 

    # 3c. Perform Matrix Multiplication (Summation)
    # [UniqueGenes x OldGenes] %*% [OldGenes x Cells] = [UniqueGenes x Cells]
    # --- SAFETY CHECK ---
    # Verify alignment: The number of columns in our mapper must match rows in data
    if (ncol(agg_mat) != nrow(counts_filtered)) {
        stop("CRITICAL ERROR: Dimension mismatch. The aggregation matrix does not align with the count matrix.")
    } else {
        message(glue("Validation Passed: Aggregating {ncol(agg_mat)} original gene rows into {nrow(agg_mat)} unique gene rows."))
        clean_counts <- agg_mat %*% counts_filtered
    }

    # STEP 4: Normalize (CPM + Log2)
    # 4a. Scaling Factors (using original library sizes)
    scaling_factors <- 1e6 / total_counts

    # 4b. Calculate CPM (Sparse Multiplication)
    # Multiply each column by its specific scaling factor
    cpm_sparse <- clean_counts %*% Matrix::Diagonal(x = scaling_factors)

    # 4c. Log2 Transform (Safe for Sparse Matrices)
    # log1p(x) / log(2) is mathematically equal to log2(x+1)
    # but it preserves zeros (sparsity), saving RAM.
    final_matrix <- log1p(cpm_sparse) / log(2)
    colnames(final_matrix) <- colnames(counts_filtered)
    # rownames(final_matrix) <- toupper(rownames(final_matrix))

    # Final Check
    message("Final Matrix Stats:")
    print(dim(final_matrix))
    print(final_matrix[1:5, 1:5]) # Preview corner

    write.csv(as.matrix(final_matrix), output_file, quote = FALSE)
  
  }
}

```

## Execute scFEA

Running scFEA script — **submits SLURM jobs**; runs only when `run_analysis <- TRUE` in `setup-display`.

```{r}
#| label: cell-scfea-sbatch
#| eval: !expr run_analysis

tissues <- c("AMY", "HYP")
for (tissue in tissues) {
  cmd <- glue(
      "sbatch {wkdir}/notebooks/shell_scripts/run_scFEA.sh",
      " --input_matrix {scFEA_input_path}/snRNASeq/snRNASeq_log2cpm_subset_{tissue}.csv")
  print(cmd)
  shell_do(cmd)
}


# sh /resnick/groups/mthomson/jboktor/WILDRxSPF_brains/notebooks/shell_scripts/run_scFEA_dynamic.sh \
# -i "/resnick/groups/mthomson/jboktor/WILDRxSPF_brains/data/input/scFEA/snRNASeq" \
# -o "/resnick/groups/mthomson/jboktor/WILDRxSPF_brains/data/interim/scFEA/snRNASeq" \
# -t "snRNASeq_log2cpm_AMY_CS20230722_CLAS_03.csv"

class_scfea_inputs <- list.files(glue("{scFEA_input_path}/snRNASeq"), pattern = "_CLAS_")
# input <- class_scfea_inputs[1]
for (input in class_scfea_inputs) {
  cmd <- glue(
    "sbatch {wkdir}/notebooks/shell_scripts/run_scFEA_dynamic.sh",
    " -i {scFEA_input_path}/snRNASeq",
    " -o {scFEA_output_path}",
    " -t '{input}'"
  )
  print(cmd)
  shell_do(cmd)
}


```

Checking results for completion


```{r}
#| eval: !expr run_analysis

# list tissue specific cell types with at least 100 cells
class_counts <- seurat_obj@meta.data %>% 
  group_by(class_label, class_name, tissue) %>% 
  tally() %>% 
  filter(n > 100) %>% 
  mutate(scfea_filename = glue("{scFEA_output_path}/flux_snRNASeq_log2cpm_{tissue}_{class_label}_run3.csv")) %>% 
  mutate(scfea_filename_exists = file.exists(scfea_filename)) %>% 
  glimpse()

class_counts$scfea_filename_exists %>% table()

```


## Analyzing scFEA results

```{r}
#| eval: !expr run_analysis

celltype_droplist <- readRDS(glue("{data_path}/seurat/snRNASeq/celltype_droplist_2025-12-30.rds"))

metadata_metabolic <- seurat_obj@meta.data %>% 
  as.data.frame() %>% 
  dplyr::select(bc_wells, group, tissue,
  # class_label, class_name, subclass_label, subclass_name, 
  supertype_label, supertype_name) %>% 
  left_join(celltype_droplist, by = c("supertype_label" = "celltype", "tissue")) %>%
  filter(is.na(dropme)) %>%
  distinct() %>% 
  glimpse()

# mouse_module_info
load("/resnick/groups/MazmanianLab/jboktor/git/scFEA/input/mouse_module_info.RData")

super_module_meta <- tribble(
  ~SM_id, ~SM_name,
  1,  "Glycolysis + TCA cycle",
  2,  "Serine Metabolism",
  3,  "Pentose phosphate",
  4,  "Fatty Acids Metabolism/Synthesis",
  5,  "Aspartate Metabolism",
  6,  "Beta-Alanine Metabolism",
  7,  "Propionyl-CoA Metabolism",
  8,  "Glutamate Metabolism",
  9,  "Leucine + Valine + Isoleucine",
  10, "Urea Cycle",
  11, "Spermine Metabolism",
  12, "Transporters",
  13, "Hyaluronic acid synthesis",
  14, "Glycogen synthesis",
  15, "Glycosaminoglycan synthesis",
  16, "N-linked glycan synthesis",
  17, "O-linked glycan synthesis",
  18, "Sialic acid synthesis",
  19, "Glycan synthesis",
  20, "Purine synthesis",
  21, "Pyrimidine synthesis",
  22, "Steroid hormone synthesis"
)
mouse_modules <- mouse_module_info %>%
  as.data.frame() %>%
  mutate(SM_id = as.numeric(SM_id)) %>% 
  left_join(super_module_meta) %>% 
  glimpse()
# mouse_modules %>% 
#   write.csv(glue("{data_path}/scFEA/mouse_modules_2026-06-08.csv"), row.names = FALSE)

flux_paths <- list.files(glue("{scFEA_output_path}"), full.names = TRUE) %>% 
  keep(~ grepl("flux", .) & grepl("_CLAS_", .))

# T-tests: compare two groups (e.g. SPF vs WildR) per supertype_name and module (M_1 .. M_171).
# Works for any tissue (AMY, HYP, etc.) — group names come from flux_df$group.
# Safe t-test: return NA when data are constant or only one group.
safe_ttest <- function(value, group) {
  out <- tryCatch(
    t.test(value ~ group),
    error = function(e) list(
      statistic = NA_real_, p.value = NA_real_,
      estimate = c(NA_real_, NA_real_)  # generic; extraction uses [1] and [2]
    )
  )
  out
}

# Looping through flux data for statistical testing
flux_ttest_results_df <- data.frame()

for (flux_path in flux_paths) {
  message(glue("Processing {flux_path}"))
  flux_df <- read.csv(flux_path) %>% 
    dplyr::rename(bc_wells = X) %>% 
    left_join(metadata_metabolic, by = "bc_wells") %>% 
    filter(!is.na(supertype_label))

  flux_ttest_results <- flux_df %>%
    pivot_longer(
      cols = matches("^M_\\d+"),
      names_to = "module",
      values_to = "value"
    ) %>%
    group_by(supertype_name, module) %>%
    filter(n_distinct(group) == 2L) %>%
    summarise(
      t_result = list(safe_ttest(value, group)),
      .groups = "drop"
    ) %>%
    mutate(
      statistic = map_dbl(t_result, "statistic"),
      p_value   = map_dbl(t_result, "p.value"),
      # Means are in alphabetical order by group name (e.g. SPF_* then WildR_*)
      mean_1 = map_dbl(t_result, ~ .x$estimate[1]),
      mean_2 = map_dbl(t_result, ~ .x$estimate[2]),
      group_1 = map_chr(t_result, ~ { n <- names(.x$estimate); if (length(n) >= 1L) sub("^mean in group ", "", n[1]) else NA_character_ }),
      group_2 = map_chr(t_result, ~ { n <- names(.x$estimate); if (length(n) >= 2L) sub("^mean in group ", "", n[2]) else NA_character_ })
    ) %>%
    mutate(
      mean_diff = mean_1 - mean_2,
      # log2FC for heatmaps: symmetric, standard. eps avoids log(0) / div by 0.
      log2FC = log2((mean_1 + 1e-10) / (mean_2 + 1e-10))
    ) %>%
    select(supertype_name, module, group_1, group_2, statistic, p_value, mean_1, mean_2, mean_diff, log2FC)

  flux_ttest_results %>% glimpse()
  flux_ttest_results_df %<>% bind_rows(flux_ttest_results)

}

flux_ttest_results_df %>% glimpse()
# saveRDS(flux_ttest_results_df, glue("{data_path}/scFEA/snRNASeq_flux_ttest_results_df_2026-02-06.rds"))

# Fast bulk Wilcox via presto: all modules per supertype in one C++ pass.
# ~10^4–10^5× faster than per-module permutation (coin); asymptotic, not perm-based.
run_flux_wilcox_presto <- function(flux_df) {
  flux_df %>%
    dplyr::select(supertype_name, group, dplyr::matches("^M_\\d+")) %>%
    filter(!is.na(supertype_name), !is.na(group)) %>%
    group_by(supertype_name) %>%
    group_modify(\(.x, .y) {
      groups <- droplevels(factor(.x$group))
      if (dplyr::n_distinct(groups) != 2L) {
        return(tibble::tibble())
      }

      group_levels <- sort(levels(groups))
      module_mat <- .x %>%
        dplyr::select(dplyr::matches("^M_\\d+")) %>%
        as.matrix() %>%
        t()
      if (ncol(module_mat) < 3L) {
        return(tibble::tibble())
      }

      presto_res <- tryCatch(
        presto::wilcoxauc(module_mat, groups, verbose = FALSE),
        error = function(e) NULL
      )
      if (is.null(presto_res) || nrow(presto_res) == 0L) {
        return(tibble::tibble())
      }

      mean_tbl <- .x %>%
        pivot_longer(
          cols = dplyr::matches("^M_\\d+"),
          names_to = "module",
          values_to = "value"
        ) %>%
        filter(is.finite(value)) %>%
        group_by(module, group) %>%
        summarise(mean_value = mean(value), .groups = "drop") %>%
        tidyr::pivot_wider(
          names_from = group,
          values_from = mean_value,
          names_prefix = "mean_"
        )

      presto_res %>%
        dplyr::filter(group == group_levels[2]) %>%
        dplyr::rename(module = feature, p_value = pval) %>%
        left_join(mean_tbl, by = "module") %>%
        mutate(
          group_1 = group_levels[1],
          group_2 = group_levels[2],
          mean_1 = .data[[paste0("mean_", group_levels[1])]],
          mean_2 = .data[[paste0("mean_", group_levels[2])]],
          mean_diff = mean_1 - mean_2,
          log2FC = log2((mean_1 + 1e-10) / (mean_2 + 1e-10))
        ) %>%
        dplyr::select(
          module, group_1, group_2, statistic, p_value,
          mean_1, mean_2, mean_diff, log2FC
        )
    }) %>%
    ungroup()
}

flux_perm_wilcox_results_df <- data.frame()

for (flux_path in flux_paths) {
  message(glue("Processing (presto Wilcox) {flux_path}"))
  flux_df <- read.csv(flux_path) %>%
    dplyr::rename(bc_wells = X) %>%
    left_join(metadata_metabolic, by = "bc_wells") %>%
    filter(!is.na(supertype_label))

  flux_perm_wilcox_results <- run_flux_wilcox_presto(flux_df)
  flux_perm_wilcox_results %>% glimpse()
  flux_perm_wilcox_results_df %<>% bind_rows(flux_perm_wilcox_results)
}

flux_perm_wilcox_results_df %>% glimpse()
# saveRDS(flux_perm_wilcox_results_df, 
#   glue("{data_path}/scFEA/snRNASeq_flux_perm_wilcox_results_df_2026-06-08.rds")
#   )

```


Visualizing T-test results


```{r}
#| eval: !expr run_analysis

cell_label_meta <- seurat_obj@meta.data %>% 
  as.data.frame() %>% 
  dplyr::select(class_label, class_name, subclass_label, subclass_name,supertype_label, supertype_name) %>% 
  distinct() %>% 
  glimpse()

# mouse_module_info
load("/resnick/groups/mthomson/jboktor/git/scFEA/input/mouse_module_info.RData")
mouse_modules <- mouse_module_info %>% as.data.frame() %>% 
  dplyr::rename(module = M_id) %>% 
  glimpse()

flux_ttest_results_df <- readRDS(glue("{data_path}/scFEA/snRNASeq_flux_ttest_results_df_2026-02-06.rds"))
flux_ttest_results_df %<>% 
  left_join(cell_label_meta) %>% 
  left_join(mouse_modules) %>% 
  mutate(tissue = str_after_last(group_1, "_")) %>% 
  glimpse()

flux_ttest_results_df %>% glimpse()
p_flux_heatmap <- flux_ttest_results_df %>% 
  filter(p_value < 0.05) %>% 
  ggplot(aes(x = module, y = supertype_name, fill = log2FC)) +
  geom_tile() +
  theme_bw() +
  facet_grid(rows = vars(tissue), cols = vars(SM_id), scales = "free", space = "free") +
  scico::scale_fill_scico(palette = 'bam', na.value = "white") +
  labs(x = "Module", y = "Supertype", fill = "Log2FC") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggsave(
  glue("{fig_path}/scFEA/snRNASeq_flux_heatmap_2026-02-06.png"),
  p_flux_heatmap,
  width = 20,
  height = 20
)

flux_ttest_results_df %>% 
  filter(p_value < 0.05) %>% 
  arrange(p_value) %>% 
  write_csv(glue("{data_path}/scFEA/snRNASeq_flux_ttest_results_df_2026-02-06.csv"))
  View()
  glimpse()

```


```{r}
#| eval: !expr run_analysis

flux_ttest_results_df <- read_csv(glue("{data_path}/scFEA/snRNASeq_flux_ttest_results_df_2026-02-06.csv")) %>% 
  glimpse()

flux_ttest_results_df %>% 
  filter(supertype_name == "1160 Astro-NT NN_2") %>% 
  View()

flux_ttest_results_df %>% 
  filter(supertype_name == "1160 Astro-NT NN_2") %>% 
  # filter(p_value < 0.001) %>% 
  group_by(SM_id, supertype_name) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n))

flux_ttest_results_df %>% 
  filter(supertype_name == "1160 Astro-NT NN_2") %>% 
  filter(SM_id == "21") %>% 
  View()

```


Visualizing Wilcoxon results

```{r}
#| eval: !expr run_analysis

cell_label_meta <- seurat_obj@meta.data %>% 
  as.data.frame() %>% 
  dplyr::select(class_label, class_name, subclass_label, subclass_name,supertype_label, supertype_name) %>% 
  distinct() %>% 
  glimpse()

# mouse_module_info
load("/resnick/groups/MazmanianLab/jboktor/git/scFEA/input/mouse_module_info.RData")
mouse_modules <- mouse_module_info %>% as.data.frame() %>% 
  dplyr::rename(module = M_id) %>% 
  mutate(SM_id = as.numeric(SM_id)) %>% 
  left_join(super_module_meta) %>% 
  glimpse()

flux_perm_wilcox_results_df <- readRDS(
  glue("{data_path}/scFEA/snRNASeq_flux_perm_wilcox_results_df_2026-06-08.rds"))
flux_perm_wilcox_results_df %<>% 
  left_join(cell_label_meta) %>% 
  left_join(mouse_modules) %>% 
  mutate(tissue = str_after_last(group_1, "_")) %>% 
  glimpse()

flux_perm_wilcox_results_df %>% glimpse()
p_flux_heatmap <- flux_perm_wilcox_results_df %>% 
  filter(p_value < 0.05) %>% 
  ggplot(aes(x = module, y = supertype_name, fill = log2FC)) +
  geom_tile() +
  theme_bw() +
  facet_grid(rows = vars(tissue), cols = vars(SM_id), scales = "free", space = "free") +
  scico::scale_fill_scico(palette = 'bam', na.value = "white") +
  labs(x = "Module", y = "Supertype", fill = "Log2FC") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

ggsave(
  glue("{fig_path}/scFEA/snRNASeq_flux_perm_wilcox_heatmap_2026-06-08.png"),
  p_flux_heatmap,
  width = 20,
  height = 20
)

flux_perm_wilcox_results_df %>% 
  filter(p_value < 0.05) %>% 
  arrange(p_value) %>% 
  write_csv(glue("{data_path}/scFEA/snRNASeq_flux_perm_wilcox_results_df_2026-06-08.csv"))

```


# correlating Cmss1/Lars2 levels with flux of spe