---
title: "Metabolic Flux Analysis"
editor: source
author: "Joe Boktor"
date: '2026-01-24'
format: 
  html:
    self-contained: true
    code-fold: true
    code-tools: true
    fig-align: right
    fig-height: 8
    fig-width: 14 
---

```{r}
library(Seurat)
library(Matrix)
library(irlba)
library(scico)
library(tidyverse)
library(patchwork)
library(glue)
library(presto)
library(hdf5r)
library(tidyseurat)
library(Augur)
library(batchtools)
library(tictoc)
library(future)
library(strex)
library(cowplot)
library(aplot)
library(SingleCellExperiment)
library(BiocParallel)
library(ggside)
theme_set(theme_bw())

home_dir <- "/resnick/groups/MazmanianLab/jboktor"
wkdir <- glue("{home_dir}/WILDRxSPF_brains")
data_path <- glue("{wkdir}/data/interim")
fig_path <- glue("{wkdir}/figures")
abca_color_pal <- readRDS(glue("{wkdir}/data/interim/abca_color_pal.rds"))
scFEA_gitpath <- glue("{home_dir}/git/scFEA")
scFEA_input_path <- glue("{wkdir}/data/input/scFEA")
scFEA_output_path <- glue("{wkdir}/data/interim/scFEA/snRNASeq")
# dir.create(scFEA_output_path, showWarnings = FALSE, recursive = TRUE)

source(glue("{wkdir}/notebooks/R_scripts/_misc_functions.R"))

# Set memory limit for future
options(future.globals.maxSize = 200000 * 1024^2)  # Set to ~200GB
# Use more memory-efficient serialization
options(future.serialization.progressr = FALSE)
options(future.serialization.progressr.enable = FALSE)

# future::plan(multisession, workers = 3)
# future::plan(sequential)
```


## Analysis Prep

Loading Seurat object

```{r}
seurat_obj <- readRDS( 
    glue("{data_path}/seurat/snRNASeq/",
    "seurat_obj_onlyparsefilters_celltyped_2025-12-27.rds")
    )
seurat_obj
# Subset the seurat object to only keep cells that pass the QC filters
seurat_obj <- seurat_obj %>% subset(subset = keep_cell)
seurat_obj@meta.data %>% glimpse()

allen_annot_df <- seurat_obj@meta.data %>% 
  select(class_name, class_label, subclass_name, 
    subclass_label, supertype_name, supertype_label) %>% 
  distinct() %>% 
  glimpse()

allen_annot_long <- allen_annot_df %>% 
  pivot_longer(
    cols = everything(),
    names_to = c("cell_type_level", ".value"),
    names_pattern = "(class|subclass|supertype)_(name|label)"
  ) %>% 
  dplyr::rename(cell_type_name = name, cell_type_label = label) %>% 
  distinct() %>% 
  glimpse()

```


Git repo for scFEA is cloned into this path `scFEA_gitpath`.


Save a cellxgene matrix for each mouse sample separately, For each sample matrix:

1. CPM normalize, log2 transform
2. Filter out specific genes in metabolic modules
3. Save matrix in our data path



```{r}
# Load Metabolic Gene List (Clean names, e.g., "Hk2", "G6pd")
metabolic_genes <- read.delim(glue("{scFEA_input_path}/scFEA.mouse.genes.txt"), header = FALSE) %>% 
  pull(V1)

# generate_cpm_matrix <- function(seurat_obj, tissue) {
tissues <- c("AMY", "HYP")
# ts <- "AMY"
for (ts in tissues) {
  # STEP 1: Get Raw Data & Calculate Library Size
  seurat_ts <- seurat_obj %>% subset(subset = tissue == ts)
  # get a list of celltypes for this tissue
  celltype_counts <- seurat_ts@meta.data$class_label %>% table()
  celltype_labels <- names(celltype_counts)[celltype_counts > 100]
  
  for (celltype in celltype_labels) {
    output_file <- glue("{scFEA_input_path}/snRNASeq/snRNASeq_log2cpm_{ts}_{celltype}.csv")
    if (file.exists(output_file)) {
      message(glue("Skipping {celltype} in {ts} because it already exists"))
      next
    }
    message(glue("Processing {celltype} in {ts}"))
    seurat_celltype <- seurat_ts %>% subset(subset = class_label == celltype)
    # check if the celltype has at least 100 cells
    if (ncol(seurat_celltype) < 100) {
      message(glue("Number of cells in {celltype} in {ts}: {ncol(seurat_celltype)}"))
      message(glue("Skipping {celltype} in {ts} because it has less than 100 cells"))
      next
    }
    counts <- Seurat::GetAssayData(seurat_celltype, layer = "counts")
    total_counts <- Matrix::colSums(counts)

    # STEP 2: Filter for Metabolic Genes
    metabolism_gene_lgl <- seurat_celltype[["RNA"]][[]]$gene_name %in% metabolic_genes
    counts_filtered <- counts[metabolism_gene_lgl, ]
    # Sanity Check
    message(glue("Original Genes: {nrow(counts)}"))
    message(glue("Filtered Metabolic Genes: {nrow(counts_filtered)}"))
    message(glue("Genes Removed: {sum(!metabolism_gene_lgl)}"))


    # STEP 3: Aggregate Isoforms (The Matrix Multiplication Method)
    # 3a. Clean Gene Names (Remove suffixes like -GRCm39)
    clean_genes <- strex::str_before_first(rownames(counts_filtered), "-GRCm39")

    # 3b. create the Transformation Matrix
    # as.factor sorts levels alphabetically -> Rows of agg_mat will be sorted!
    clean_genes_factor <- as.factor(clean_genes)
    agg_mat <- as(clean_genes_factor, "sparseMatrix") 

    # 3c. Perform Matrix Multiplication (Summation)
    # [UniqueGenes x OldGenes] %*% [OldGenes x Cells] = [UniqueGenes x Cells]
    # --- SAFETY CHECK ---
    # Verify alignment: The number of columns in our mapper must match rows in data
    if (ncol(agg_mat) != nrow(counts_filtered)) {
        stop("CRITICAL ERROR: Dimension mismatch. The aggregation matrix does not align with the count matrix.")
    } else {
        message(glue("Validation Passed: Aggregating {ncol(agg_mat)} original gene rows into {nrow(agg_mat)} unique gene rows."))
        clean_counts <- agg_mat %*% counts_filtered
    }

    # STEP 4: Normalize (CPM + Log2)
    # 4a. Scaling Factors (using original library sizes)
    scaling_factors <- 1e6 / total_counts

    # 4b. Calculate CPM (Sparse Multiplication)
    # Multiply each column by its specific scaling factor
    cpm_sparse <- clean_counts %*% Matrix::Diagonal(x = scaling_factors)

    # 4c. Log2 Transform (Safe for Sparse Matrices)
    # log1p(x) / log(2) is mathematically equal to log2(x+1)
    # but it preserves zeros (sparsity), saving RAM.
    final_matrix <- log1p(cpm_sparse) / log(2)
    colnames(final_matrix) <- colnames(counts_filtered)
    # rownames(final_matrix) <- toupper(rownames(final_matrix))

    # Final Check
    message("Final Matrix Stats:")
    print(dim(final_matrix))
    print(final_matrix[1:5, 1:5]) # Preview corner

    write.csv(as.matrix(final_matrix), output_file, quote = FALSE)
  
  }
}

# tissues <- seurat_obj@meta.data$tissue %>% unique()
# for (tissue in tissues) {
#   tic()
#   message(glue("Processing {tissue}"))
#   generate_cpm_matrix(seurat_obj, tissue)
#   toc()
# }

```

## Execute scFEA

Running scFEA script
```{r}
tissues <- c("AMY", "HYP")
for (tissue in tissues) {
  cmd <- glue(
      "sbatch {wkdir}/notebooks/shell_scripts/run_scFEA.sh",
      " --input_matrix {scFEA_input_path}/snRNASeq/snRNASeq_log2cpm_subset_{tissue}.csv")
  print(cmd)
  shell_do(cmd)
}


sh /resnick/groups/MazmanianLab/jboktor/WILDRxSPF_brains/notebooks/shell_scripts/run_scFEA_dynamic.sh \
-i "/resnick/groups/MazmanianLab/jboktor/WILDRxSPF_brains/data/input/scFEA/snRNASeq" \
-o "/resnick/groups/MazmanianLab/jboktor/WILDRxSPF_brains/data/interim/scFEA/snRNASeq" \
-t "snRNASeq_log2cpm_AMY_CS20230722_CLAS_03.csv"

class_scfea_inputs <- list.files(glue("{scFEA_input_path}/snRNASeq"), pattern = "_CLAS_")
# input <- class_scfea_inputs[1]
for (input in class_scfea_inputs) {
  cmd <- glue(
    "sbatch {wkdir}/notebooks/shell_scripts/run_scFEA_dynamic.sh",
    " -i {scFEA_input_path}/snRNASeq",
    " -o {scFEA_output_path}",
    " -t '{input}'"
  )
  print(cmd)
  shell_do(cmd)
}


```

Checking results for completion


```{r}
# list tissue specific cell types with at least 100 cells
class_counts <- seurat_obj@meta.data %>% 
  group_by(class_label, class_name, tissue) %>% 
  tally() %>% 
  filter(n > 100) %>% 
  mutate(scfea_filename = glue("{scFEA_output_path}/flux_snRNASeq_log2cpm_{tissue}_{class_label}_run3.csv")) %>% 
  mutate(scfea_filename_exists = file.exists(scfea_filename)) %>% 
  glimpse()

class_counts$scfea_filename_exists %>% table()

```


## Analyzing scFEA results

```{r}
metadata_metabolic <- seurat_obj@meta.data %>% 
  as.data.frame() %>% 
  dplyr::select(bc_wells, group, class_label, class_name, subclass_label, subclass_name, supertype_label, supertype_name) %>% 
  distinct() %>% 
  glimpse()

# mouse_module_info
load("/resnick/groups/MazmanianLab/jboktor/git/scFEA/input/mouse_module_info.RData")
mouse_modules <- mouse_module_info %>% as.data.frame() %>% glimpse

flux_paths <- list.files(glue("{scFEA_output_path}"), full.names = TRUE) %>% 
  keep(~ grepl("flux", .) & grepl("_CLAS_", .))

flux_df <- read.csv(flux_paths[1]) %>% 
  dplyr::rename(bc_wells = X) %>% 
  left_join(metadata_metabolic, by = "bc_wells") %>% 
  glimpse()

flux_df %>% 
  ggplot(aes(x=M123, y=group )) +
  geom_point() +
  theme_bw() +
  facet_wrap(~class_name, scales = "free")

# add celltype names
# add microbiome status of celltype





```
