---
title: "pySCENIC Gene Regulatory Networks in WildR vs SPF Brain"
subtitle: "snRNA-seq — pilot subsample + tissue×class production (Flux-style)"
date: today
author: "Joe Boktor"
categories: [snRNA-seq, SCENIC, Gene Regulatory Networks]
format:
  html:
    theme:
      - cosmo
      - ../quarto-aesthetic-kit/theme/aesthetic.scss
    include-in-header: ../quarto-aesthetic-kit/theme/fonts.html
    grid:
      body-width: 1280px
      sidebar-width: 180px
      margin-width: 240px
    toc: true
    toc-location: right
    toc-title: "On this page"
    page-layout: article
    code-fold: true
    code-tools: true
    code-copy: true
    embed-resources: true
    fig-cap-location: bottom
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup-display
#| include: false

# Normal render/preview: FALSE — no Seurat load, no loom export, no sbatch.
# Set TRUE intentionally to export class looms and/or submit SLURM jobs (appendix).
run_analysis <- FALSE

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(knitr)
  library(glue)
  library(stringr)
  library(jsonlite)
})

wkdir <- "/resnick/groups/mthomson/jboktor/WILDRxSPF_brains"
fig_dir <- file.path(wkdir, "figures/SCENIC/snRNASeq")
proc_dir <- file.path(wkdir, "data/processed/scenic")
agg_fig_dir <- file.path(wkdir, "figures/SCENIC/snRNASeq_by_class/_aggregate")
agg_proc_dir <- file.path(wkdir, "data/processed/scenic_by_class")
interim_dir <- file.path(wkdir, "data/interim/scenic")
input_db <- file.path(wkdir, "data/input/scenic/cisTarget")

fig_rel <- function(name) file.path("../figures/SCENIC/snRNASeq", name)
agg_fig_rel <- function(name) file.path("../figures/SCENIC/snRNASeq_by_class/_aggregate", name)
proc_rel <- function(name) file.path("../data/processed/scenic", name)
agg_proc_rel <- function(name) file.path("../data/processed/scenic_by_class", name)

has_fig <- function(name) file.exists(file.path(fig_dir, name))
has_agg_fig <- function(name) file.exists(file.path(agg_fig_dir, name))
has_proc <- function(...) {
  paths <- file.path(proc_dir, c(...))
  any(file.exists(paths))
}

# Prefer dated CSVs if present, else latest match
latest_csv <- function(pattern, dir = proc_dir) {
  hits <- sort(list.files(dir, pattern = pattern, full.names = TRUE))
  if (!length(hits)) return(NA_character_)
  hits[length(hits)]
}

diff_csv <- latest_csv("^diff_regulons_WildR_vs_SPF_.*\\.csv$")
rss_csv <- latest_csv("^regulon_rss_.*\\.csv$")
summary_json <- file.path(proc_dir, "run_summary.json")
conclusions_txt <- file.path(proc_dir, "conclusions.txt")

run_summary <- if (file.exists(summary_json)) {
  jsonlite::fromJSON(summary_json)
} else {
  NULL
}

n_cells <- if (!is.null(run_summary)) run_summary$n_cells else NA
n_genes <- if (!is.null(run_summary)) run_summary$n_genes else NA
n_regulons <- if (!is.null(run_summary)) run_summary$n_regulons else NA
n_sig <- if (!is.null(run_summary)) run_summary$n_sig_diff else NA

# Class-split aggregation (production)
agg_summary_json <- file.path(agg_proc_dir, "aggregate_summary.json")
agg_summary <- if (file.exists(agg_summary_json)) jsonlite::fromJSON(agg_summary_json) else NULL
hit_counts_csv <- file.path(agg_proc_dir, "hit_counts_by_split.csv")
tf_recurrence_csv <- file.path(agg_proc_dir, "tf_hit_recurrence.csv")
top_hits_csv <- file.path(agg_proc_dir, "top_hits_padj05.csv")
agg_concl_txt <- file.path(agg_proc_dir, "conclusions.txt")
run_status_csv <- file.path(agg_proc_dir, "run_status.csv")
```

# pySCENIC Gene Regulatory Networks in WildR vs SPF Brain

## Overview

Production pySCENIC runs use **raw UMI counts** from Seurat `RNA` `counts`, one GRN per **tissue × Allen class** (Flux-style; both microbiome groups in each loom). WildR vs SPF contrasts are Wilcoxon tests on AUCell within each split. Cross-class comparisons use **TF hit lists**, not raw AUC (regulons are inferred separately per split).

AMY `01 IT-ET Glut` (~87k cells) and the two whole-tissue jobs are still running and are **excluded** from the aggregation below.

```{r}
#| label: display-stats
#| echo: false
#| results: asis

if (is.null(agg_summary)) {
  cat("::: {.callout-warning}\n")
  cat("**Class-split aggregates not found.** Run ")
  cat("`notebooks/R_scripts/aggregate_scenic_class_results.R`.\n")
  cat(":::\n")
} else {
  pending <- paste(agg_summary$pending_stems, collapse = ", ")
  cat("::: {.stat-row}\n")
  cat("::: {.stat}\n")
  cat(glue("[{agg_summary$n_completed}/{agg_summary$n_completed + agg_summary$n_pending}]{{.stat-value}}\n"))
  cat("[Class splits done]{.stat-label}\n")
  cat(":::\n")
  cat("::: {.stat}\n")
  cat(glue("[{format(agg_summary$n_sig_tests, big.mark=',')}]{{.stat-value}}\n"))
  cat("[Sig. regulon tests]{.stat-label}\n")
  cat(":::\n")
  cat("::: {.stat}\n")
  cat(glue("[{agg_summary$n_splits_with_hits}]{{.stat-value}}\n"))
  cat("[Splits with hits]{.stat-label}\n")
  cat(":::\n")
  cat("::: {.stat}\n")
  cat(glue("[{agg_summary$n_pending}]{{.stat-value}}\n"))
  cat("[Pending (excluded)]{.stat-label}\n")
  cat(":::\n")
  cat(":::\n")
  if (nzchar(pending)) {
    cat("::: {.callout-note}\n")
    cat(glue("Pending / excluded from aggregation: `{pending}`.\n"))
    cat(":::\n")
  }
}
```

## Background

Microbiome status (WildR vs SPF) reshapes brain gene expression in this cohort (see DEG notebooks). SCENIC asks which **transcription-factor regulons** change activity. Gene symbols were stripped of the Parse `-GRCm39` suffix to match cisTarget TF lists. Input matrices are **raw UMIs** (not the Seurat `data` layer).

## Results — class-split production

### Run completeness

::: {.figure-card .fig-wide}
```{r}
#| label: display-agg-status
#| echo: false
#| results: asis
if (has_agg_fig("run_status.png")) {
  cat(sprintf("![](%s)\n", agg_fig_rel("run_status.png")))
} else {
  cat("*Aggregate figure pending.*\n")
}
```

::: {.fig-caption}
[**Figure P1.**]{.fig-num} Class-split run status by cell count. Orange = pending (excluded from aggregation).
:::
:::

### Significant regulons per class

::: {.figure-card .fig-wide}
```{r}
#| label: display-agg-nsig
#| echo: false
#| results: asis
if (has_agg_fig("n_sig_by_class.png")) {
  cat(sprintf("![](%s)\n", agg_fig_rel("n_sig_by_class.png")))
} else {
  cat("*Aggregate figure pending.*\n")
}
```

::: {.fig-caption}
[**Figure P2.**]{.fig-num} Number of BH-significant (padj < 0.05) WildR vs SPF regulons per completed tissue × class split.
:::

::: {.takeaways}
**Key takeaways**

- Glial classes dominate hit counts: **Astro-Epen** and **OPC-Oligo** in both tissues; HYP **CNU-LGE GABA** also high.
- Several neuronal classes show moderate hit sets (e.g. CNU-HYa, CTX-MGE/CGE in AMY; HY GABA/Glut in HYP).
- Small / rare classes often have **zero** significant regulons — underpowered, not necessarily null biology.
:::
:::

```{r}
#| label: display-hit-counts-table
#| echo: false
#| results: asis

if (file.exists(hit_counts_csv)) {
  hc <- read_csv(hit_counts_csv, show_col_types = FALSE) %>%
    filter(status == "completed") %>%
    arrange(desc(n_sig)) %>%
    transmute(
      tissue, class = class_name,
      n_cells = format(n_cells, big.mark = ","),
      n_regulons = n_regulons_tested,
      n_sig, WildR_up = n_WildR_up, SPF_up = n_SPF_up
    )
  cat(knitr::kable(hc, caption = "Significant regulons per completed class split.") |> paste(collapse = "\n"))
  cat("\n\n")
}
```

### Direction of shifts

::: {.figure-card .fig-wide}
```{r}
#| label: display-agg-direction
#| echo: false
#| results: asis
if (has_agg_fig("sig_direction_by_class.png")) {
  cat(sprintf("![](%s)\n", agg_fig_rel("sig_direction_by_class.png")))
} else {
  cat("*Aggregate figure pending.*\n")
}
```

::: {.fig-caption}
[**Figure P3.**]{.fig-num} Among significant regulons, counts higher in WildR vs higher in SPF (mean AUCell).
:::
:::

### Combined volcano

::: {.figure-card .fig-wide}
```{r}
#| label: display-agg-volcano
#| echo: false
#| results: asis
if (has_agg_fig("volcano_all_splits.png")) {
  cat(sprintf("![](%s)\n", agg_fig_rel("volcano_all_splits.png")))
} else {
  cat("*Aggregate figure pending.*\n")
}
```

::: {.fig-caption}
[**Figure P4.**]{.fig-num} All within-split WildR − SPF regulon tests from completed class GRNs (faceted by tissue).
:::
:::

### Recurrent TFs across classes

::: {.figure-card .fig-wide}
```{r}
#| label: display-agg-tf-recurrence
#| echo: false
#| results: asis
if (has_agg_fig("tf_recurrence.png")) {
  cat(sprintf("![](%s)\n", agg_fig_rel("tf_recurrence.png")))
}
if (has_agg_fig("tf_hit_heatmap.png")) {
  cat(sprintf("\n![](%s)\n", agg_fig_rel("tf_hit_heatmap.png")))
}
```

::: {.fig-caption}
[**Figure P5.**]{.fig-num} TFs whose regulons are significant in ≥2 class splits (hit-list recurrence), and presence/direction heatmap across splits.
:::

::: {.takeaways}
**Key takeaways**

- Most recurrent TFs include **Rfx3**, **Bclaf1**, **Klf12**, **Bptf**, **Mllt10**, **Nfia**, **Pbx1** — candidates for microbiome-linked GRN state across multiple classes.
- **Mllt10** / **Ube2k** hits skew WildR-up across splits; others are mixed — direction is class-dependent.
- Treat recurrence as hypothesis generation; confirm against DEG TF lists and (when finished) tissue-wide shared-regulon AUCs.
:::
:::

```{r}
#| label: display-top-hits-table
#| echo: false
#| results: asis

if (file.exists(top_hits_csv)) {
  top <- read_csv(top_hits_csv, show_col_types = FALSE) %>%
    arrange(padj, desc(abs(delta))) %>%
    transmute(
      tissue, class = class_name, tf, regulon,
      delta, padj, direction
    ) %>%
    slice_head(n = 40)
  cat("### Top differential regulons (padj < 0.05)\n\n")
  cat(knitr::kable(top, caption = "Top 40 class-split hits by padj.") |> paste(collapse = "\n"))
  cat("\n\nFull table: [`", agg_proc_rel("top_hits_padj05.csv"), "`]{.path}\n", sep = "")
}
```

```{r}
#| label: display-agg-conclusions
#| echo: false
#| results: asis

if (file.exists(agg_concl_txt)) {
  cat("::: {.callout-note}\n")
  cat("**Auto-summary (class aggregation)**\n\n```\n")
  cat(paste(readLines(agg_concl_txt), collapse = "\n"))
  cat("\n```\n:::\n")
}
```

## Results — pilot subsample (reference)

Early ~5k-cell pilot used to validate the pipeline (gene naming, RSS identity TFs). Production claims should use the class-split section above.

### Pilot subsample composition

::: {.figure-card .fig-wide}
```{r}
#| label: display-composition
#| echo: false
#| results: asis
if (has_fig("subsample_composition.png")) {
  cat(sprintf("![](%s)\n", fig_rel("subsample_composition.png")))
} else {
  cat("*Figure pending — pipeline not finished.*\n")
}
```

::: {.fig-caption}
[**Figure 1.**]{.fig-num} Stratified pilot subsample composition by Allen class, tissue (AMY / HYP), and microbiome (WildR / SPF). Sampling capped per microbiome × tissue × subclass, then globally to ~5,000 cells.
:::

::: {.takeaways}
**Key takeaways**

- Both tissues and microbiome groups are represented, so WildR vs SPF contrasts are not driven by missing strata.
- Large glutamatergic / GABAergic and glial classes dominate the subsample, matching cohort composition.
- This is a **pilot**; rare subclasses remain underpowered for per-type regulon tests.
:::

[Source: cell `cell-run-pipeline`](#cell-run-pipeline){.provenance}
:::

Artifacts: intermediates in [data/interim/scenic/]{.path}; cisTarget DBs in [data/input/scenic/cisTarget/]{.path}.

### Regulon activity embedding

::: {.figure-card .fig-wide}
```{r}
#| label: display-umap
#| echo: false
#| results: asis
if (has_fig("scenic_umap_feature.png")) {
  cat(sprintf("![](%s)\n", fig_rel("scenic_umap_feature.png")))
} else {
  cat("*Figure pending — pipeline not finished.*\n")
}
```

::: {.fig-caption}
[**Figure 2.**]{.fig-num} UMAP of AUCell regulon-activity space. Top: cells colored by Allen subclass and by microbiome. Bottom: FeaturePlots for the four most variable regulons in the pilot.
:::

::: {.takeaways}
**Key takeaways**

- Regulon AUC embeddings largely recover **cell-type identity**, as expected for SCENIC.
- Microbiome labels **mix within** subclasses rather than forming separate clouds — consistent with state shifts inside types, not wholesale retyping.
- High-variance regulons that track subclasses are candidate identity TFs; those that vary by microbiome within type are candidate state TFs.
:::

[Source: cell `cell-run-pipeline`](#cell-run-pipeline){.provenance}
:::

### Top variable regulons across cell types

::: {.figure-card .fig-wide}
```{r}
#| label: display-heatmap
#| echo: false
#| results: asis
if (has_fig("regulon_heatmap_top50.png")) {
  cat(sprintf("![](%s)\n", fig_rel("regulon_heatmap_top50.png")))
} else {
  cat("*Figure pending — pipeline not finished.*\n")
}
```

::: {.fig-caption}
[**Figure 3.**]{.fig-num} Pseudobulk scaled AUCell scores for the 50 most variable regulons, columns grouped by microbiome (SPF / WildR) and annotated by tissue.
:::

::: {.takeaways}
**Key takeaways**

- Column structure is dominated by **cell-type programs** (blocks of regulons co-active in related subclasses).
- Within-type SPF vs WildR column pairs can reveal microbiome-linked TF programs when present.
- Use this heatmap to pick regulons for FeaturePlots and for overlap with DEG hits.
:::

[Source: cell `cell-run-pipeline`](#cell-run-pipeline){.provenance}
:::

::: {.figure-card .fig-wide}
```{r}
#| label: display-dotplot
#| echo: false
#| results: asis
if (has_fig("regulon_dotplot.png")) {
  cat(sprintf("![](%s)\n", fig_rel("regulon_dotplot.png")))
} else {
  cat("*Figure pending — pipeline not finished.*\n")
}
```

::: {.fig-caption}
[**Figure 4.**]{.fig-num} Dot plot of the top 25 variable regulons across the 12 largest subclasses, split by microbiome within each subclass.
:::

::: {.takeaways}
**Key takeaways**

- Dot size/color summarize how broadly and strongly each regulon is active in a subclass × microbiome group.
- Systematic WildR–SPF differences within a subclass nominate TFs for follow-up with DEG and RSS results.
:::

[Source: cell `cell-run-pipeline`](#cell-run-pipeline){.provenance}
:::

### Cell-type–specific regulons (RSS)

::: {.figure-card .fig-wide}
```{r}
#| label: display-rss
#| echo: false
#| results: asis
if (has_fig("regulon_rss.png")) {
  cat(sprintf("![](%s)\n", fig_rel("regulon_rss.png")))
} else {
  cat("*Figure pending — pipeline not finished.*\n")
}
```

::: {.fig-caption}
[**Figure 5.**]{.fig-num} Regulon Specificity Score (RSS) — top five regulons per major subclass. RSS uses a Jensen–Shannon–based score relating AUCell profiles to cell-type membership (SCENIC protocol style).
:::

::: {.takeaways}
**Key takeaways**

- RSS recovers expected identity TFs: **Myb / Irx5** (ependymal), **Runx1 / Irf5 / Irf8 / Ikzf1** (microglia), **Sox10** (oligo).
- That is a strong sanity check that motif pruning + AUCell worked on this Parse gene set after `-GRCm39` stripping.
- Identity regulons that later show microbiome ΔAUC would be identity × state candidates; none are FDR-significant in this pilot.
- Full RSS table: [data/processed/scenic/regulon_rss_2026-08-08.csv]{.path}.
:::

[Source: cell `cell-run-pipeline`](#cell-run-pipeline){.provenance}
:::

### Differential regulon activity: WildR vs SPF

::: {.figure-card .fig-wide}
```{r}
#| label: display-volcano
#| echo: false
#| results: asis
if (has_fig("diff_regulon_volcano.png")) {
  cat(sprintf("![](%s)\n", fig_rel("diff_regulon_volcano.png")))
} else {
  cat("*Figure pending — pipeline not finished.*\n")
}
```

::: {.fig-caption}
[**Figure 6.**]{.fig-num} Volcano of Wilcoxon tests of AUCell scores (WildR − SPF) within tissue × subclass strata (minimum 10 cells per group). Highlighted points: padj < 0.05 and large |ΔAUC|.
:::

::: {.takeaways}
**Key takeaways**

- After BH correction across ~9k tissue × subclass × regulon tests, **no regulon reaches padj < 0.05** in this 5k-cell pilot.
- Nominal top hits (e.g. Tcf12, Zfp148 in HYP subclasses) exist but are underpowered; treat as hypotheses only.
- Positive ΔAUC = higher mean activity in WildR; negative = higher in SPF.
- Scale cell numbers (or lock regulons and score the full object) before concluding microbiome GRN effects are absent.
:::

[Source: cell `cell-run-pipeline`](#cell-run-pipeline){.provenance}
:::

```{r}
#| label: display-diff-table
#| echo: false
#| results: asis

if (!is.na(diff_csv) && file.exists(diff_csv)) {
  diff <- read_csv(diff_csv, show_col_types = FALSE)
  top <- diff %>%
    filter(!is.na(padj), padj < 0.05) %>%
    arrange(padj) %>%
    transmute(
      tissue, subclass_name, regulon,
      n_SPF, n_WildR,
      delta = round(delta, 4),
      p = signif(p, 3),
      padj = signif(padj, 3)
    ) %>%
    slice_head(n = 25)
  cat("### Top differential regulons\n\n")
  if (nrow(top) == 0) {
    cat("No regulon tests reached padj < 0.05 in this pilot subsample. ")
    cat("Expand `target_n_cells` / `max_cells_per_group` before concluding absence of GRN effects.\n")
  } else {
    cat(knitr::kable(top, caption = "Top WildR vs SPF differential regulons (padj < 0.05).") |> paste(collapse = "\n"))
    cat("\n\nFull table: [`", proc_rel(basename(diff_csv)), "`]{.path}\n", sep = "")
  }
} else {
  cat("::: {.callout-note}\nDifferential table not written yet.\n:::\n")
}
```

### Conclusions

```{r}
#| label: display-conclusions
#| echo: false
#| results: asis

if (file.exists(conclusions_txt)) {
  cat("::: {.callout-note}\n")
  cat("**Auto-summary from pipeline**\n\n")
  cat("```\n")
  cat(paste(readLines(conclusions_txt), collapse = "\n"))
  cat("\n```\n")
  cat(":::\n")
}
```

**Interpretation checklist (standard pySCENIC read)**

1. **AUC UMAP** — subclass structure present → identity regulons dominate (expected). Look for microbiome mixing *within* subclass as state change.
2. **RSS** — recover known lineage TFs; if not, check gene naming / DB mismatch / underpowered subsample.
3. **WildR − SPF volcano** — few pilot hits is common; scale up before claiming “no GRN effect.”
4. **DEG overlap** — TFs that are DE *and* whose regulons change AUC are the strongest mechanistic candidates.

Processed Seurat with SCENIC assay: [data/processed/scenic/seurat_with_scenic_auc.rds]{.path}.

## Production design: two parallel workflows

We run **two** pySCENIC production tracks from the same raw UMI counts:

1. **Class split (Flux-style)** — one GRN per tissue × Allen `class_label` (~100+ cells; both microbiome groups in the loom). Best for within-type WildR vs SPF state tests; jobs are smaller and parallel.
2. **Tissue split** — one GRN for all AMY nuclei (~199k) and one for all HYP (~190k). Enables within-tissue comparisons across cell types on a **shared** regulon vocabulary. If these finish in reasonable walltime, prefer them for cross-class questions; otherwise fall back to hit-list comparison across class GRNs.

| | Class split | Tissue split |
|--|-------------|--------------|
| Unit | tissue × `class_label` | whole AMY / whole HYP |
| WildR vs SPF | within class | overall + within-class on shared regulons |
| Cross-class AUC | not comparable across jobs | comparable within tissue |
| Compute | many medium jobs | 2 large jobs (600G / 7d requested) |

### Input matrix (verified)

pySCENIC / GRNBoost2 expects **raw counts** (tree models; official guidance: no need to normalize). We export Seurat `RNA` **`counts`** only — not `data`.

Checks on `seurat_obj_onlyparsefilters_celltyped_2025-12-27.rds` (`keep_cell`):

- `counts` nonzeros are **100% integer**, non-negative; library sizes look like UMIs (sample median ~13.5k).
- `data` is a transformed layer (non-integer; not used).
- Gene names `Symbol-GRCm39` are stripped to bare symbols for cisTarget TF overlap.

Export refuses to write a loom if the counts layer fails the integer check.

# Code & Provenance {.code-appendix}

## Launch: export + submit both workflows

Preferred one-shot (compute node): `run_scenic_export_and_submit.sh` exports class + tissue looms, then `sbatch`s class jobs (default 24 CPU / 300G / 3d) and tissue jobs (32 CPU / 600G / 7d).

```{r}
#| label: cell-scenic-launch
#| eval: !expr run_analysis

source(file.path(wkdir, "notebooks/R_scripts/_misc_functions.R"))
cmd <- glue("sbatch {wkdir}/notebooks/shell_scripts/run_scenic_export_and_submit.sh")
print(cmd)
shell_do(cmd)
```

```{r}
#| label: cell-scenic-check
#| eval: !expr run_analysis

input_dir <- file.path(wkdir, "data/input/scenic/snRNASeq")
man_class <- file.path(input_dir, "manifest_class.csv")
man_tissue <- file.path(input_dir, "manifest_tissue.csv")
if (file.exists(man_class)) {
  message("Class manifest:")
  print(read_csv(man_class, show_col_types = FALSE))
}
if (file.exists(man_tissue)) {
  message("Tissue manifest:")
  print(read_csv(man_tissue, show_col_types = FALSE))
}
done_c <- list.files(file.path(wkdir, "data/interim/scenic/by_class"),
                     pattern = "run_summary\\.json$", recursive = TRUE)
done_t <- list.files(file.path(wkdir, "data/interim/scenic/by_tissue"),
                     pattern = "run_summary\\.json$", recursive = TRUE)
message("Finished class summaries: ", length(done_c),
        "; tissue summaries: ", length(done_t))
```

## Optional: pilot end-to-end (subsample)

```{r}
#| label: cell-run-pipeline
#| eval: !expr run_analysis

# Pilot only — production uses class + tissue SLURM jobs above.
script <- file.path(wkdir, "notebooks/R_scripts/run_scenic_pipeline.R")
stopifnot(file.exists(script))
status <- system2(
  "/resnick/groups/MazmanianLab/jboktor/software/miniforge3/envs/spatialomics/bin/Rscript",
  args = script
)
stopifnot(status == 0)
```

### How to regenerate

```bash
# One-shot: export raw-count looms + submit class and tissue pySCENIC jobs
sbatch notebooks/shell_scripts/run_scenic_export_and_submit.sh

# Aggregate completed class splits (skips pending, e.g. AMY CLAS_01)
/resnick/groups/MazmanianLab/jboktor/software/miniforge3/envs/spatialomics/bin/Rscript \
  notebooks/R_scripts/aggregate_scenic_class_results.R
```

### Path conventions

| Role | Path |
|------|------|
| cisTarget DBs / TF list | `data/input/scenic/cisTarget/` |
| Looms + manifests | `data/input/scenic/snRNASeq/` |
| Class interim | `data/interim/scenic/by_class/<stem>/` |
| Tissue interim | `data/interim/scenic/by_tissue/<stem>/` |
| Class aggregate tables | `data/processed/scenic_by_class/` |
| Class aggregate figures | `figures/SCENIC/snRNASeq_by_class/_aggregate/` |
| Class / tissue per-split figures | `figures/SCENIC/snRNASeq_by_{class,tissue}/` |
| SLURM logs / job list | `.cluster_runs/scenic/` |
| Pilot (subsample) | `data/interim/scenic/`, `data/processed/scenic/`, `figures/SCENIC/snRNASeq/` |
